\section{Implementation Details}
\label{app:details}

This appendix records all implementation details necessary for reproducibility: training hyperparameters, RL configuration, prompt templates, and library versions.

%------------------------------------------------------------------------------
\subsection{Libraries and Software}
\label{app:details:libraries}

\begin{table}[h]
\centering
\small
\begin{tabular}{lll}
\toprule
\textbf{Package} & \textbf{Version} & \textbf{Role} \\
\midrule
PyTorch & 2.8.0+cu126 & Training backend \\
VERL & 0.7.1 & RL training framework \\
vLLM & 0.10.2 & Rollout inference engine (within VERL's $\le 0.12$ range) \\
Ray & 2.55.1 & Distributed orchestration \\
Transformers & 4.56.2 (pinned) & Model loading \& tokenization \\
lc0 & 0.32.1 & Chess specialist engine (UCI binary) \\
python-chess & 1.11.2 & UCI client / FEN handling \\
FastAPI / Uvicorn & 0.136 / 0.46 & HTTP wrapper for the engine \\
CUDA toolkit & 12.6 & Training (lc0 binary built against 12.4) \\
\bottomrule
\end{tabular}
\caption{Key software dependencies.}
\label{tab:libraries}
\end{table}

\paragraph{Version-coupling notes.}
The vLLM/PyTorch/Transformers versions are tightly coupled and cannot be
changed independently. vLLM~0.10.2 is the highest release inside VERL~0.7.1's
supported range ($\geq 0.8.5$, $\leq 0.12.0$) that ships the
\texttt{qwen3\_next.py} model file (linear-attention/full-attention hybrid
support). Its pre-built CUDA wheel calls the two-argument form of
\texttt{c10::cuda::SetDevice}, which is only present in PyTorch~$\geq 2.8$.
Transformers~$\geq 5$ removes \texttt{all\_special\_tokens\_extended} from
the tokenizer base class, which vLLM~0.10.2's tokenizer wrapper still
references; we pin to 4.56.2.

%------------------------------------------------------------------------------
\subsection{Model}
\label{app:details:model}

The base LLM is \texttt{Qwen/Qwen3-4B}: a 4\,B-parameter decoder-only Qwen3
dense model with the standard \texttt{Qwen3ForCausalLM} architecture,
served by vLLM with the Hermes tool-call parser. We initially targeted
\texttt{Qwen/Qwen3.5-4B} (a hybrid linear-attention/full-attention model
declared as \texttt{Qwen3\_5ForConditionalGeneration}) but no
VERL-supported vLLM release recognises that architecture string, so the
toy task uses Qwen3-4B as the base.

The projector module, latent-state token count $K{=}16$, and adapter
geometry are described in the main paper.

%------------------------------------------------------------------------------
\subsection{Training Hyperparameters}
\label{app:details:hparams}

% TODO: fill in per-stage hyperparams (Stage 1 projector pretraining, Stage 2 GRPO fine-tuning)

\begin{table}[h]
\centering
\small
\begin{tabular}{lll}
\toprule
\textbf{Hyperparameter} & \textbf{Stage 1} & \textbf{Stage 2 (GRPO)} \\
\midrule
Learning rate & -- & -- \\
Batch size & -- & -- \\
Epochs & -- & -- \\
KL coefficient & -- & -- \\
$K$ (latent tokens) & 32 & 32 \\
\bottomrule
\end{tabular}
\caption{Training hyperparameters.}
\label{tab:hparams}
\end{table}

%------------------------------------------------------------------------------
\subsection{RL Configuration (VERL + vLLM)}
\label{app:details:rl}

Reinforcement learning uses GRPO as the advantage estimator, implemented via the VERL framework~\cite{} with an external vLLM rollout server.

% TODO: rollout temperature, top-p, response length, reward shaping details

%------------------------------------------------------------------------------
\subsection{Chess Specialist: lc0 + BT4}
\label{app:details:lc0}

The chess specialist agent is Leela Chess Zero (\texttt{lc0}) v0.32.1 running the BT4 transformer network. The LLM accesses it through an HTTP wrapper (\texttt{lc0\_server}); no in-process binding into the LLM training stack — the wrapper is a process boundary.

\paragraph{Engine binary.}
Built from upstream source (commit \texttt{v0.32.1}) with \texttt{gcc 11.4}, \texttt{meson} + \texttt{ninja}, CUDA toolkit 12.4. Build invocation: \texttt{./build.sh release -Dgtest=false -Ddnnl=false}. Both \texttt{cuda} (fp32) and \texttt{cuda-fp16} backends are produced; we run \texttt{cuda-fp16} on A100 (SM 8.0). The binary is dynamically linked, lives at \texttt{/dev/shm/somesh/lc0\_src/build/release/lc0}, and exposes a UCI interface on stdin/stdout. lc0's CUDA backend uses cuBLAS only — no cuDNN dependency.

\paragraph{Network.}
\texttt{BT4-1024x15x32h-swa-6147500-policytune-332.pb.gz} from the lczero ``contrib networks'' bucket: a 1024-dim, 15-block, 32-head BT4 transformer, multihead format (policy / value / WDL / moves-left), saved at training step 6{,}147{,}500 after a SWA pass and a policy-tuning fine-tune (\texttt{policytune-332}). On disk: 365\,MB compressed; resident GPU footprint at our settings: $\sim$\,2.9\,GB (weights in fp16 plus cuBLAS workspace).

\paragraph{Tool interface.}
The engine is exposed via FastAPI on a pool of eight servers, one per A100 GPU, bound to \texttt{http://localhost:7100}--\texttt{7107}. The \texttt{LcOClient} in the inference harness round-robins requests across the pool via \texttt{itertools.cycle}, providing 8$\times$ parallel throughput for batch evaluation. Three endpoints are shared across all instances (Table~\ref{tab:lc0-api}). The LLM never speaks UCI directly; it issues JSON over HTTP, mediated by the wrapper. The same FEN-based interface generalises to any UCI engine, so swapping in Stockfish or Leela contempt nets later requires no LLM-side change.

\begin{table}[h]
\centering
\small
\begin{tabular}{lll}
\toprule
\textbf{Endpoint} & \textbf{Body} & \textbf{Returns} \\
\midrule
\texttt{GET /health} & --- & \texttt{\{status, engine\_id, backend, weights\}} \\
\texttt{POST /analyze} & \texttt{\{fen, nodes?, movetime\_ms?, multipv?\}} & \texttt{\{bestmove, multipv:[\{score\_cp, mate, depth, nodes, time\_ms, pv\}]\}} \\
\texttt{POST /policy} & \texttt{\{fen, nodes?\}} & \texttt{\{value\_root, moves:[\{move, P, N, Q, WL, D, V, U, S, M\}]\}} \\
\bottomrule
\end{tabular}
\caption{HTTP interface to \texttt{lc0\_server}. \texttt{/policy} parses lc0's \texttt{VerboseMoveStats} info-string lines: \texttt{P} is the NN prior in $[0,1]$, \texttt{V} the per-child single-eval network value in $[-1, +1]$, \texttt{Q} the running average action-value, \texttt{N} the visit count. With \texttt{nodes} $\geq$ \#\,legal\,moves, every root child is expanded exactly once and \texttt{P}/\texttt{V} reduce to a pure forward pass.}
\label{tab:lc0-api}
\end{table}

\paragraph{Engine flags --- overrides versus defaults.}
We deliberately leave search-behaviour flags at the upstream defaults so the engine matches publicly reproducible Leela behaviour; we override only the resource and observability flags driven by GPU-sharing (the engine co-resides on a single A100 with the RL rollout vLLM, dp=8). Both groups are listed in Table~\ref{tab:lc0-flags} for full reproducibility. Authoritative descriptions and defaults are taken from the lczero documentation~\cite{lc0flags}.

\begin{table}[h]
\centering
\small
\begin{tabular}{llll}
\toprule
\textbf{Parameter} & \textbf{Upstream default} & \textbf{Our value} & \textbf{Rationale} \\
\midrule
\multicolumn{4}{l}{\emph{Overridden (resources / observability)}} \\
\midrule
\texttt{Backend}              & \texttt{cuda-auto} & \texttt{cuda-fp16} & Explicit fp16 on Ampere \\
\texttt{WeightsFile}          & \texttt{<autodiscover>} & BT4 path & Pinned weights \\
\texttt{Threads}              & \texttt{0} (backend) & \texttt{2} & Two CPU workers feeding the GPU batch \\
\texttt{MinibatchSize}        & \texttt{0} (backend) & \texttt{128} & Throughput sweet spot on A100 (Table~\ref{tab:lc0-bench}) \\
\texttt{MaxPrefetch}          & \texttt{32} & \texttt{32} & Upstream default --- memory is flat, no reason to throttle \\
\texttt{NNCacheSize}          & \texttt{2{,}000{,}000} & \texttt{200{,}000} & Cap host RAM; FEN-cache hits dominate at low N \\
\texttt{VerboseMoveStats}     & \texttt{false} & \texttt{true} & Required for the \texttt{/policy} parser \\
\texttt{MultiPV}              & \texttt{1} & per-request & Set per call by python-chess \\
\midrule
\multicolumn{4}{l}{\emph{Held at upstream defaults (search behaviour)}} \\
\midrule
\texttt{CPuct}                  & \texttt{1.75}    & \multicolumn{2}{l}{cpuct\_init for PUCT} \\
\texttt{CPuctBase}              & \texttt{38739}   & \multicolumn{2}{l}{cpuct\_base for PUCT growth} \\
\texttt{CPuctFactor}            & \texttt{3.89}    & \multicolumn{2}{l}{cpuct growth multiplier} \\
\texttt{FpuStrategy}            & \texttt{reduction}& \multicolumn{2}{l}{First-Play Urgency mode} \\
\texttt{FpuValue}               & \texttt{0.33}    & \multicolumn{2}{l}{FPU offset for unvisited nodes} \\
\texttt{PolicyTemperature}      & \texttt{1.36}    & \multicolumn{2}{l}{Policy softmax temperature} \\
\texttt{OutOfOrderEval}         & \texttt{true}    & \multicolumn{2}{l}{Cache / terminal short-circuit} \\
\texttt{MaxCollisionEvents}     & \texttt{917}     & \multicolumn{2}{l}{Collision events per batch} \\
\texttt{MaxCollisionVisits}     & \texttt{80000}   & \multicolumn{2}{l}{Total collision visits per batch} \\
\texttt{MaxConcurrentSearchers} & \texttt{1}       & \multicolumn{2}{l}{Single batch-gathering worker} \\
\texttt{ContemptMode}           & \texttt{play}    & \multicolumn{2}{l}{Symmetric WDL} \\
\texttt{ScoreType}              & \texttt{WDL\_mu} & \multicolumn{2}{l}{Score = $100\!\cdot\!Q$, not centipawns} \\
\texttt{UCI\_ShowWDL}           & \texttt{false}   & \multicolumn{2}{l}{Off (we recover WDL via /policy)} \\
\texttt{UCI\_ShowMovesLeft}     & \texttt{false}   & \multicolumn{2}{l}{Off} \\
\bottomrule
\end{tabular}
\caption{lc0 configuration. All search-behaviour flags are at upstream defaults so results match unmodified Leela.}
\label{tab:lc0-flags}
\end{table}

\paragraph{Throughput / batch profile (A100, BT4, cuda-fp16).}
Resident GPU memory is essentially \emph{flat} across batch sizes (Table~\ref{tab:lc0-bench}); the cuBLAS workspace plus fp16 weights dominate, and per-position activations are negligible at this scale. Hence \texttt{MinibatchSize=128} is set as the default --- it gives 3$\times$ the throughput of \texttt{MinibatchSize=8} for free.

\begin{table}[h]
\centering
\small
\begin{tabular}{rrrr}
\toprule
\textbf{Batch} & \textbf{Peak GPU mem} & \textbf{Throughput} & \textbf{ms/batch} \\
\midrule
1    & 2983 MiB & 366 nps    & 2.7   \\
8    & 2983 MiB & 2{,}413 nps   & 3.3   \\
32   & 2983 MiB & 5{,}335 nps   & 6.0   \\
64   & 2983 MiB & 6{,}839 nps   & 9.4   \\
128  & 2983 MiB & 7{,}319 nps   & 17.5  \\
256  & 2983 MiB & 7{,}754 nps   & 33.0  \\
512  & 2983 MiB & 8{,}051 nps   & 63.6  \\
1024 & 2983 MiB & 8{,}283 nps   & 123.6 \\
\bottomrule
\end{tabular}
\caption{\texttt{lc0 backendbench} sweep on A100-80GB, BT4 + cuda-fp16. Memory is constant at $\sim$\,3 GiB regardless of batch size; throughput plateaus near batch 128--256. Co-residing engines on the same A100 only need to leave $\sim$\,4\,GiB ($\sim$\,5\%) headroom per GPU.}
\label{tab:lc0-bench}
\end{table}

\paragraph{Wrapper configuration.}
Environment variables, with defaults: \texttt{LC0\_BIN} (engine path), \texttt{LC0\_NET} (\texttt{.pb.gz} weights), \texttt{LC0\_BACKEND} (\texttt{cuda-fp16}), \texttt{LC0\_THREADS} (\texttt{2}), \texttt{LC0\_MINIBATCH} (\texttt{128}), \texttt{LC0\_MAX\_PREFETCH} (\texttt{32}), \texttt{LC0\_NNCACHE} (\texttt{200000}), \texttt{LC0\_HOST} (\texttt{0.0.0.0}), \texttt{LC0\_PORT} (\texttt{7100}). The multi-GPU deployment script \texttt{start\_lc0\_servers.sh} launches eight instances in parallel, assigning \texttt{CUDA\_VISIBLE\_DEVICES=}$i$ and \texttt{LC0\_PORT=}$7100+i$ for $i\in\{0,\ldots,7\}$, then health-checks all eight before returning. Each wrapper instance holds a single \texttt{Lc0Engine}; HTTP-level concurrency within each process is serialised by an internal lock because UCI is a single-conversation protocol. Both \texttt{/analyze} and \texttt{/policy} accept an optional \texttt{moves} field (UCI-string list) applied on top of the FEN, so BT4's seven history planes are filled with the real game rather than synthesised; this matters for raw-policy queries mid-game (the \texttt{HistoryFill=fen\_only} default otherwise repeats the current FEN). \texttt{/policy} also accepts a per-call \texttt{policy\_temperature} override; note this only affects fresh evaluations because lc0's NN cache memoises the post-temperature priors per FEN.

\paragraph{Strongest raw policy (top-$k$ from the NN with no search).}
For $\arg\max_m P(m)$ to come from the unmodified BT4 forward pass: (i) call \texttt{/policy} with \texttt{nodes=1} and the full \texttt{moves} list so the history planes are real; (ii) pick the move with highest \texttt{P}. Temperature does not affect the argmax. For top-$k$ ranking, temperature also does not change the order. For probability-faithful sampling against the NN's training distribution, set \texttt{policy\_temperature=1.0} (the upstream default of \texttt{1.36} is tuned for MCTS exploration, not for honest sampling). The \texttt{policytune-332} fine-tune was specifically designed to sharpen the raw policy head, so this path is what the network is most optimised for.

\paragraph{Score conventions.}
With \texttt{ScoreType=WDL\_mu}, the centipawn field reported by lc0 is $100\!\cdot\!Q$ where $Q\in[-1,+1]$ is the WDL-mean utility from the side to move's perspective. For the experiments in \S\ref{sec:experiments} we report this as-is; it is \emph{not} a Stockfish-style centipawn evaluation. The \texttt{value\_root} field on \texttt{/policy} is the visit-weighted $Q$ across root children (a proxy for $V$ at the root), with $V$ available directly for each child once expanded.

\paragraph{Strength.}
We do not have a direct measurement of CCRL Elo for the \texttt{policytune-332} BT4 net. As a rough orientation, the BT4 family at long time control sits at $\sim$\,3500--3700 Elo on a single A100; at $1{,}000$ nodes per move (a typical setting for the \texttt{/analyze} endpoint) the network plus MCTS is $\sim$\,3300; at $1$ node (raw top policy) it is $\sim$\,2400--2600 -- a strong IM / weak GM ``intuition'' move. A calibration script (\texttt{lc0\_server/scripts/bench\_elo.py}) is provided to measure this against Stockfish at fixed depth on demand.

\paragraph{Latent extraction (forward-looking).}
The UCI interface is sufficient for LLAMIA-Bench (the \emph{verbalized}-access baseline) but does \emph{not} expose hidden activations. The \emph{latent state internalization} path (\S\ref{sec:methodology}) requires a separate forward-pass shim that loads the same BT4 \texttt{.pb.gz} weights into PyTorch and emits the post-attention residual stream we project into $K{=}16$ tokens; this is implemented out of process and is not part of \texttt{lc0\_server}.

%------------------------------------------------------------------------------
\subsection{Chess Analyst Agent (Inference Harness)}
\label{app:details:agent}

% ── Shared palette & style presets ───────────────────────────────────────────
\definecolor{agentslate}{HTML}{2d3748}
\definecolor{agentteal}{HTML}{234e52}
\definecolor{agentmint}{HTML}{edf7f5}
\definecolor{agentlav}{HTML}{f3eeff}
\definecolor{agentrose}{HTML}{fff0f0}
\definecolor{agentgold}{HTML}{7c5c00}
\definecolor{agentgoldbg}{HTML}{fffbec}

\tcbset{
  % ── floating-title base ────────────────────────────────────────────
  agentbase/.style={
    enhanced, breakable,
    arc=4pt, outer arc=4pt,
    left=9pt, right=9pt, top=9pt, bottom=9pt,
    before skip=11pt, after skip=11pt,
    attach boxed title to top left={xshift=10pt, yshift=-\tcboxedtitleheight/2},
    boxed title style={arc=3pt, outer arc=3pt, size=small, left=5pt, right=5pt},
    drop shadow={color=black!18, opacity=0.55, shadow xshift=1.5pt, shadow yshift=-1.5pt},
  },
  % ── Prompt / instruction boxes ─────────────────────────────────────
  promptbox/.style={
    agentbase,
    colback=agentlav, colframe=themepurple!80,
    fonttitle=\bfseries\sffamily\small\color{white},
    fontupper=\small,
    boxed title style={colback=themepurple, colframe=themepurple, arc=3pt, size=small, left=5pt, right=5pt},
  },
  % ── Tool definition boxes ──────────────────────────────────────────
  toolbox/.style={
    agentbase,
    colback=agentmint, colframe=agentteal,
    fonttitle=\bfseries\ttfamily\small\color{white},
    fontupper=\small,
    boxed title style={colback=agentteal, colframe=agentteal, arc=3pt, size=small, left=5pt, right=5pt},
  },
  % ── State / architecture boxes ─────────────────────────────────────
  statebox/.style={
    agentbase,
    colback=black!3, colframe=agentslate!70,
    fonttitle=\bfseries\sffamily\small\color{white},
    fontupper=\small,
    boxed title style={colback=agentslate, colframe=agentslate, arc=3pt, size=small, left=5pt, right=5pt},
  },
  % ── Warning / caveat boxes ─────────────────────────────────────────
  warnbox/.style={
    agentbase,
    colback=agentrose, colframe=themered!80,
    fonttitle=\bfseries\sffamily\small\color{white},
    fontupper=\small,
    boxed title style={colback=themered, colframe=themered, arc=3pt, size=small, left=5pt, right=5pt},
  },
  % ── Note / insight boxes ───────────────────────────────────────────
  notebox/.style={
    agentbase,
    colback=agentgoldbg, colframe=agentgold!80,
    fonttitle=\bfseries\sffamily\small\color{agentgold},
    fontupper=\small,
    boxed title style={colback=agentgoldbg, colframe=agentgold, arc=3pt, size=small, left=5pt, right=5pt},
  },
}

% ── Helper: italic key--value line inside a box ──────────────────────────────
\newcommand{\toolparam}[2]{%
  \hspace*{4pt}{\ttfamily\bfseries #1}\,\texttt{:}\;\textit{#2}\\[1pt]}
\newcommand{\toolret}[1]{%
  \smallskip\noindent{\color{agentslate}\textbf{Returns:}}\;\textit{#1}}
\newcommand{\tooldesc}[1]{%
  \noindent{\color{agentslate}\textbf{Description:}}\;#1\smallskip}

\medskip
The inference harness connects the LLM to the lc0 engine via a six-tool
stateful API.  Figure~\ref{fig:LLAMIA} shows the high-level architecture;
this section records the exact state representation, system prompt, and tool
schemas used in all LLAMIA-Bench experiments.

% ─────────────────────────────────────────────────────────────────────────────
\subsubsection{Agent State}
\label{app:details:agent:state}

\begin{tcolorbox}[statebox, title={Agent State \textnormal{(\texttt{agents/state.py} — \texttt{ChessState})}},
  label={box:agentstate}]

\textbf{\textsf{Fields held in memory across tool calls:}}

\smallskip
\begin{tabular}{@{}lp{2.5cm}p{8.4cm}@{}}
  \toprule
  \textbf{Field} & \textbf{Type} & \textbf{Content} \\
  \midrule
  \texttt{board} & \texttt{chess.Board} & Full game position (Zobrist hash, castling rights, en-passant target, 50-move clock). \\
  \texttt{\_history} & \texttt{list[str]} & SAN move list from the start of the current episode. \\
  \bottomrule
\end{tabular}

\medskip
\textbf{\textsf{Derived properties (computed on demand):}}

\smallskip
\begin{tabular}{@{}lp{10.5cm}@{}}
  \toprule
  \textbf{Property} & \textbf{Value} \\
  \midrule
  \texttt{fen} & \texttt{board.fen()} --- passed to every lc0 API call \\
  \texttt{info()} & FEN, ASCII board, turn, move number, in-check flag, legal SAN list (up to 24), last six moves \\
  \bottomrule
\end{tabular}

\smallskip
\noindent The \texttt{ascii\_board} field emits \texttt{str(chess.Board)}: an
8\,$\times$\,8 grid with white pieces uppercase and black lowercase.  This
representation lets the LLM read piece placements visually instead of
decoding the FEN string, which removes the need for an explicit
``examine the board'' tool round before analysis.

\medskip
\textbf{\textsf{Move parsing priority.}} UCI first (\texttt{e2e4},
\texttt{g1f3}, \ldots), then SAN (\texttt{Nf3}, \texttt{Nxd4},
\texttt{O\nobreakdash-O}) via \texttt{python-chess}.  Illegal moves return
an \texttt{\{"error": \ldots\}} dict without raising; the LLM sees the
error and can retry with a corrected move.

\smallskip
\textbf{\textsf{Persistence scope.}} A single \texttt{ChessAnalyst.run()}
call.  State is reset at the start of each query or when the LLM calls
\texttt{reset\_position}.  No cross-query memory is maintained.
\end{tcolorbox}

% ─────────────────────────────────────────────────────────────────────────────
\subsubsection{System Prompt}
\label{app:details:agent:prompt}

\begin{tcolorbox}[promptbox,
  title={$\triangleright$\; System Prompt --- Chess Analyst},
  label={box:sysprompt}]

\begin{quote}
You are a chess expert with access to lc0, a top neural network engine.
A board is already loaded --- any FEN in the user's query has been
applied for you.  Do not call \texttt{reset\_position} unless you need a
different position.

\smallskip
Tools (the board is stateful across calls):
\begin{itemize}[noitemsep, topsep=2pt, leftmargin=14pt]
  \item \texttt{get\_position} --- FEN, ASCII board, legal moves, history.
  \item \texttt{make\_move(move)} --- apply UCI (\texttt{e2e4}) or SAN
        (\texttt{Nf3}) move \emph{permanently}.
  \item \texttt{undo\_move} --- take back the last move.
  \item \texttt{reset\_position(fen)} --- switch to a different FEN.
  \item \texttt{analyze(nodes, multipv, moves=[])} --- engine search.
        Pass \texttt{moves=[X]} to analyse the position after playing
        \texttt{X} without changing state --- use this for ``what if?''
        questions instead of \texttt{make\_move\,/\,undo}.
  \item \texttt{get\_policy(nodes)} --- raw NN priors $P$ and values
        $V$ per move (fast, single forward pass).
\end{itemize}

\smallskip
Blunder analysis pattern (two calls --- issue them \emph{together} in
one response):
\begin{itemize}[noitemsep, topsep=2pt, leftmargin=14pt]
  \item \texttt{analyze()} --- engine's best move and eval for the
        current position.
  \item \texttt{analyze(moves=[\textit{candidate}])} --- eval after the
        candidate move.
\end{itemize}
The difference between the two evaluations is the centipawn loss; the
PV of the second call starts with the opponent's refutation (in SAN,
e.g.\ \texttt{Qxg5+}).

\smallskip
Efficiency: when two tool calls are independent, emit \emph{both} in
the same response.  The harness dispatches them in parallel and you
save a round-trip.
\end{quote}

\smallskip
\noindent\textbf{Design rationale.}  The prompt is compact
(${\approx}\,220$ tokens); the current \texttt{max\_model\_len=32{,}000}
provides ample headroom, but brevity keeps the instruction signal dense
and avoids over-specifying a workflow.  Three clauses encode the harness's
invariants:

\begin{itemize}[noitemsep, topsep=2pt, leftmargin=14pt]
  \item The ``do not call \texttt{reset\_position}'' clause prevents
        the agent from spending its first round verifying a FEN load
        the harness has already performed.
  \item The \texttt{moves=[\ldots]} clause teaches the no-mutate
        analysis primitive: in a single call the agent obtains the
        engine's evaluation of any hypothetical move sequence without
        mutating the board.  This collapses the otherwise standard
        \texttt{make\_move}\,$\to$\,\texttt{analyze}\,$\to$\,\texttt{undo\_move}
        sequence into one tool call.
  \item The efficiency clause encourages the model to batch
        independent calls into a single response.  Qwen3.5 honours
        this for the blunder pattern, allowing both engine queries to
        complete in one LLM round.
\end{itemize}

\noindent We deliberately do not include a verbose ``workflow'' hint of
the form ``check position $\to$ try the move $\to$ analyse
$\to$ undo $\to$ compare alternatives''.  Such hints recover an
inefficient sequence; the prompt above points the model directly at
the two-call pattern, which is the information-theoretic minimum for
blunder analysis (one position evaluation + one post-move evaluation).
\end{tcolorbox}

% ─────────────────────────────────────────────────────────────────────────────
\subsubsection{Tool Definitions}
\label{app:details:agent:tools}

The six tools below are passed to the \texttt{/v1/chat/completions} endpoint in
the standard OpenAI \texttt{tools} array; vllm injects them into Qwen3.5's
chat template as the function catalogue.

\medskip

% ── Row 1: position tools ────────────────────────────────────────────────────
\begin{tcolorbox}[toolbox, title={\texttt{get\_position}()}]
\tooldesc{Returns a snapshot of the current board state.  No parameters.}

\toolret{\texttt{\{fen, ascii\_board, turn, move\_number, in\_check,\\
\hspace*{1.2em}legal\_moves[0..23], total\_legal, recent\_moves[-6:]\}}}

\smallskip
\noindent The \texttt{ascii\_board} field is \texttt{str(chess.Board)}: an
8\,$\times$\,8 grid with white pieces uppercase and black lowercase
(e.g.\ \texttt{R N B Q K B N R} on the back rank).  This lets the model
read piece placements visually rather than decoding the FEN string.  The
legal-move list is truncated to 24 entries to bound context;
\texttt{total\_legal} reports the true count.
\end{tcolorbox}

\begin{tcolorbox}[toolbox, title={\texttt{make\_move(move: str)}}]
\tooldesc{Apply a move to the current board.  Accepts UCI or SAN
  notation.  The state is mutated permanently until \texttt{undo\_move}
  or \texttt{reset\_position} is called.}

\smallskip
\begin{tabular}{@{}lp{1.5cm}p{8.5cm}@{}}
  \toprule
  \textbf{Parameter} & \textbf{Type} & \textbf{Description} \\
  \midrule
  \texttt{move} & \texttt{str} & Move in UCI (\texttt{e2e4}) or SAN (\texttt{Nxd4}, \texttt{O\nobreakdash-O\nobreakdash-O}) notation. \\
  \bottomrule
\end{tabular}

\smallskip
\toolret{\texttt{\{move\_played, fen, turn, in\_check, checkmate\}}}

\smallskip
\noindent Illegal moves return \texttt{\{"error":\,\ldots\}} without
raising; the LLM may retry with a corrected move string.
\end{tcolorbox}

\begin{tcolorbox}[toolbox, title={\texttt{undo\_move()}}]
\tooldesc{Revert the last move applied by \texttt{make\_move}.  No parameters.
  Can be called repeatedly to step back through the variation.  Returns
  \texttt{\{"error": "No moves to undo"\}} if the stack is empty.}

\toolret{\texttt{\{fen, turn\}}}
\end{tcolorbox}

\begin{tcolorbox}[toolbox, title={\texttt{reset\_position(fen: str)} \textnormal{--- locked when FEN auto-loaded}}]
\tooldesc{Replace the current board with a new position given by a FEN
  string and clear the move history.  A FEN embedded anywhere in the
  user's query is automatically extracted by the harness and applied
  before the agent loop begins; in that case this tool is removed from
  the catalogue and the dispatcher refuses any call that survives the
  model's training prior.}

\smallskip
\begin{tabular}{@{}lp{1.5cm}p{8.5cm}@{}}
  \toprule
  \textbf{Parameter} & \textbf{Type} & \textbf{Description} \\
  \midrule
  \texttt{fen} & \texttt{str} & Any valid Forsyth--Edwards Notation string \\
  \bottomrule
\end{tabular}

\smallskip
\toolret{\texttt{\{fen, status: "ok"\}} when permitted; \texttt{\{noop, message, current\_fen\}} when locked.}

\smallskip
\noindent \textbf{Locking rationale.}  When the user's query already contains
a FEN, the harness extracts and loads it before the agent loop begins.
Without the lock, Qwen3.5 still tends to call \texttt{reset\_position}
on its first turn as a habitual ``verify the load'' step --- silently
succeeding because the FEN matches but consuming an entire round.
Removing the tool from the catalogue, plus a dispatcher-level guard
that returns a no-op message if the call survives the model's training
prior, eliminates this wasted round.
\end{tcolorbox}

\begin{tcolorbox}[toolbox,
  title={\texttt{analyze(nodes: int = 800, multipv: int = 3, moves: list[str] = [])}}]
\tooldesc{Run lc0 MCTS search via \texttt{POST /analyze}, optionally
  after applying a sequence of hypothetical moves to the FEN, without
  mutating \texttt{ChessState}.  The \texttt{moves} parameter is the
  no-mutate primitive that replaces the
  \texttt{make\_move}\,$\to$\,\texttt{analyze}\,$\to$\,\texttt{undo\_move}
  sequence with a single call.  Scores are in $100\!\cdot\!Q$ units
  (see \S\ref{app:details:lc0}), positive meaning good for the side to
  move.}

\smallskip
\begin{tabular}{@{}lp{1.8cm}p{7.2cm}@{}}
  \toprule
  \textbf{Parameter} & \textbf{Type} & \textbf{Description} \\
  \midrule
  \texttt{nodes}   & \texttt{int}       & MCTS node budget (default 800; 1\,000--2\,000 for deeper PVs). \\
  \texttt{multipv} & \texttt{int}       & Number of lines to return (default 3). \\
  \texttt{moves}   & \texttt{list[str]} & UCI or SAN moves applied client-side before analysis; state unchanged.  Returned as SAN in \texttt{applied\_moves}. \\
  \bottomrule
\end{tabular}

\smallskip
\toolret{\texttt{\{turn, bestmove, lines: ["1. <SAN-pv> score=\ldots\\
\hspace*{1.2em}depth=\ldots", \ldots], applied\_moves?\}}}

\smallskip
\noindent\textbf{PV translation to SAN.}  lc0 emits principal variations
in UCI (e.g.\ \texttt{a5g5}), which hides captures, checks, and piece
identities.  Without translation, the LLM cannot tell that \texttt{a5g5}
is in fact \texttt{Qxg5} (a queen capture), and produces explanations
consistent with the headline evaluation but inconsistent with the actual
piece interactions.  The harness therefore replays each PV on a
\texttt{python-chess} board and emits SAN, so the LLM reads the
literal \texttt{Qxg5+} and quotes it directly.  The \texttt{bestmove}
field is re-emitted in SAN under the same scheme; UCI is preserved as a
fallback for any move that fails to parse.
\end{tcolorbox}

\begin{tcolorbox}[toolbox, title={\texttt{get\_policy(nodes: int = auto)}}]
\tooldesc{Run a minimal search ($\texttt{nodes} \geq \#\text{legal-moves}+2$ by default so
  every root child is visited exactly once) and return lc0's per-move neural
  network statistics.  Much faster than \texttt{analyze} and gives raw NN
  beliefs before MCTS modifies them.}

\smallskip
\begin{tabular}{@{}lll@{}}
  \toprule
  \textbf{Parameter} & \textbf{Type} & \textbf{Description} \\
  \midrule
  \texttt{nodes} & \texttt{int} & Budget (default: auto = \(\#\text{legal}+2\)) \\
  \bottomrule
\end{tabular}

\smallskip
\toolret{\texttt{\{turn, value\_root, nodes\_searched,\\
\hspace*{1.2em}top\_moves: ["<uci> P=\% V=\ldots Q=\ldots N=\ldots", \ldots]\}}}

\smallskip
\noindent Moves are reported in UCI rather than SAN: the policy view is
typically consumed as a probability distribution, and the identifier
that is comparable across calls (and across a game) is the UCI move
rather than its position-dependent SAN.  The fields per move are:
\begin{itemize}[noitemsep, topsep=2pt, leftmargin=14pt]
  \item \textbf{P} — NN prior probability in $[0,1]$; reflects the raw policy head (post \texttt{PolicyTemperature=1.36})
  \item \textbf{V} — single-eval NN value at the child node in $[-1,+1]$
  \item \textbf{Q} — visit-averaged action-value in $[-1,+1]$ (equals V when $N{=}1$)
  \item \textbf{N} — MCTS visit count; equal to 1 for every child when nodes$=$auto
\end{itemize}
\end{tcolorbox}

% ─────────────────────────────────────────────────────────────────────────────
\subsubsection{Tool-Call Format and vLLM Compatibility}
\label{app:details:agent:format}

\begin{tcolorbox}[warnbox,
  title={$\mathbf{!}$\; Tool-Call Format — Qwen3.5 XML (not Hermes JSON)},
  label={box:toolformat}]

\textbf{Root cause.}
Qwen3.5's chat template instructs the model to generate tool calls in an
XML-like format:

\smallskip
\texttt{\phantom{xx}<tool\_call>}\\
\texttt{\phantom{xxxx}<function=\textit{name}>}\\
\texttt{\phantom{xxxxxx}<parameter=\textit{key}>\textit{value}</parameter>}\\
\texttt{\phantom{xxxx}</function>}\\
\texttt{\phantom{xx}</tool\_call>}

\medskip
\noindent vllm's \texttt{--tool-call-parser hermes} locates the
\texttt{<tool\_call>} tags correctly but expects \emph{JSON} inside; failing
to parse the XML body, it leaves \texttt{msg.tool\_calls = []} and returns the
raw XML in \texttt{msg.content}.

\medskip
\textbf{Fix (no server change required).}
\texttt{\_parse\_qwen\_tool\_calls(content)} in \texttt{agents/chess\_analyst.py}
uses two regex patterns:

\smallskip
\texttt{QWEN\_TOOL\_RE} matches \texttt{<tool\_call>{\ldots}<function=(\textbackslash w+)>({\ldots})</function>{\ldots}</tool\_call>}\\
\texttt{PARAM\_RE} matches \texttt{<parameter=(\textbackslash w+)>({\ldots})</parameter>}

\smallskip
\noindent Parameter values are coerced via \texttt{json.loads}; on failure they remain
strings.  Tool results are returned as standard \texttt{role=tool} messages —
vllm reformats them correctly into Qwen3.5's expected tool-response template
on the server side, so the model reads results correctly.

\medskip
\textbf{Consequence for future work.}  Any code that reads \texttt{msg.tool\_calls}
directly from the OpenAI client will silently drop all tool calls with this
model and server configuration.  Always use \texttt{\_parse\_qwen\_tool\_calls}
(or the equivalent) when working with Qwen3.5 via the current vllm setup.
\end{tcolorbox}

% ─────────────────────────────────────────────────────────────────────────────
\subsubsection{Inference Hyperparameters}
\label{app:details:agent:hparams}

\begin{tcolorbox}[notebox, title={Inference Configuration}]
\renewcommand{\arraystretch}{1.15}
\begin{tabular}{@{}p{3.4cm}p{2.8cm}p{8cm}@{}}
  \toprule
  \textbf{Parameter} & \textbf{Value} & \textbf{Notes} \\
  \midrule
  Model & \texttt{Qwen/Qwen3-4B} & Single-replica vLLM at \texttt{:7000}; \texttt{gpu\_memory\_utilization=0.5}, \texttt{--enforce-eager} (skips Triton/cudagraph compile). \\
  \texttt{max\_model\_len}   & 8\,192 tokens & Shared across system prompt, history, tool I/O, response. (Lower than the 32\,k upstream limit; ample for tool-call rollouts that fit in $\sim$3 rounds.) \\
  \texttt{max\_tokens}        & 700           & Per LLM call; leaves ${\approx}\,7\,500$ for history. \\
  \texttt{temperature}        & 0.0 (smoke) / 1.0 (RL) & Greedy at evaluation; T${=}1.0$ during RL rollouts per the standard tool-RL recipe. \\
  \texttt{tool\_choice}       & \texttt{"auto"} & Model decides when to stop calling tools. \\
  Tool-call parsing            & dual-path    & vLLM's Hermes parser populates \texttt{message.tool\_calls}; the harness reads that field directly and falls back to the legacy Qwen3.5 inline-XML format if empty. \\
  \texttt{enable\_thinking}   & \texttt{False} & Suppressed via \texttt{chat\_template\_kwargs}; dodges the documented Qwen3 ``plans tool but never emits'' failure mode. \\
  \texttt{MAX\_ROUNDS}        & 14            & Cap on LLM calls per query; safety bound above the typical 2--3. \\
  Read-only dispatch          & parallel      & \texttt{analyze}, \texttt{get\_policy}, \texttt{get\_position} fan out via \texttt{ThreadPoolExecutor} when batched. \\
  Mutating dispatch           & sequential    & \texttt{make\_move}, \texttt{undo\_move}, \texttt{reset\_position} preserve emission order. \\
  lc0 \texttt{nodes} (analyze)& 800 (default) & Caller may override (1\,000--2\,000 for deeper PVs). \\
  lc0 \texttt{nodes} (policy) & auto          & $\max(\#\text{legal}+2,\,8)$; one visit per child. \\
  lc0 \texttt{multipv}        & 3             & Top-3 lines; chosen to bound output tokens. \\
  \bottomrule
\end{tabular}
\end{tcolorbox}

% ─────────────────────────────────────────────────────────────────────────────
\subsubsection{Worked Example: Blunder Analysis}
\label{app:details:agent:example}

To make the data flow concrete, we trace the harness on a single query:
``\textsl{Why is \texttt{Nxd4} a blunder here, \texttt{rnbqkb1r/pp2pppp/5n2/6B1/\linebreak
2pp4/4PN2/PP3PPP/RN1QKB1R w KQkq - 0 6}?}''  The position is reached from a
Trompowsky-style move order (e.g.\ \texttt{1.\,d4 d5 2.\,Bg5 c6 3.\,Nf3 Nf6
4.\,e3 Bf5} with the colour-flipped equivalent of the diagram above); White,
to move, is tempted to capture on \texttt{d4} with the knight but loses the
\texttt{Bg5} bishop after \texttt{Qa5+ Nc3 Qxg5}.

\begin{tcolorbox}[notebox, title={Tool-call trace}]

\noindent The harness extracts and loads the FEN before any LLM round.
The agent then issues two \texttt{analyze} calls in a single response;
because both are read-only they are dispatched in parallel.

\smallskip
\begin{tabular}{@{}rlp{8.6cm}@{}}
  \toprule
  \textbf{\#} & \textbf{Call} & \textbf{Engine output (SAN principal variation)} \\
  \midrule
  1 & \texttt{analyze(multipv=3)} &
    Best: \texttt{exd4}, score $-0.07$;
    PV \texttt{exd4 Be6 Nbd2 h6 Bxf6 exf6} — equal play. \\
  2 & \texttt{analyze(moves=[Nxd4], multipv=3)} &
    Best for Black: \texttt{Qa5+}, score $+7.89$;
    PV \texttt{Qa5+ Nc3 Qxg5 Bxc4 Qxg2 Rf1} — Black wins the \texttt{Bg5}
    bishop. \\
  \bottomrule
\end{tabular}

\medskip
\noindent The two evaluations differ by ${\approx}\,8$ pawns from White's
perspective.  The PV from call~2 starts with the literal SAN
\texttt{Qxg5}, so the LLM reads the refutation directly rather than
inferring it from a UCI string that hides the capture.

\medskip
\noindent\textbf{Final answer (verbatim, abbreviated):}

\smallskip
\begin{tcolorbox}[colback=black!4, colframe=black!25, sharp corners, boxrule=0.4pt,
  left=8pt, right=8pt, top=6pt, bottom=6pt, fontupper=\small]
``\texttt{Nxd4} is a blunder because it allows Black to win material with a
devastating check.\,[\ldots] \textbf{\texttt{Qa5+}}\,[\ldots] then
\textbf{\texttt{Qxg5}}, capturing the White Bishop on \texttt{g5}.\,[\ldots]
White should have captured with the pawn (\textbf{\texttt{exd4}}) instead.''
\end{tcolorbox}

\medskip
\noindent\textbf{Engine verdict.}  \texttt{Nxd4} drops the \texttt{Bg5}
bishop to \texttt{Qa5+}: a single move that simultaneously checks the
white king on the \texttt{a5\,$\to$\,e1} diagonal and attacks the bishop
along rank \texttt{5}.  After any block, \texttt{Qxg5} wins the bishop.
The correct move is \texttt{exd4} (${\approx}\,0$ eval).  The full
exchange takes 2 \texttt{lc0} calls and 2 LLM rounds end-to-end.
\end{tcolorbox}

%------------------------------------------------------------------------------
\subsection{Toy Task: Puzzle Popularity \& ELO}
\label{app:details:toytask}

We validated the verbalized-tool-use baseline on a single LLAMIA-Bench
dimension before scaling to the full suite: predicting Lichess puzzle
\emph{popularity} (an integer in $[-100, 100]$, the upvote score) and
\emph{ELO} (the Lichess puzzle difficulty rating, roughly $400$--$3000$).

\paragraph{Dataset.}
Source: \texttt{ssingh22/llamia-chess-data}, configuration
\texttt{behavioural\_cloning}, filtered to rows with
\texttt{class == "type\_Puzzle"}. Each kept row carries a starting FEN in
\texttt{state[0]} and a popularity/ELO target parsed from the third
\texttt{gpt} conversation turn (matching the regex
\verb|popularity is (-?\d+) and the ELO is (\d+)\b|).
We materialise 2\,000 train / 200 val rows in parquet format with columns
\texttt{prompt} (a 2-message system+user list), \texttt{reward\_model}
(\texttt{style="puzzle\_popularity\_elo"} and \texttt{ground\_truth} =
\texttt{\{popularity, elo\}}), \texttt{fen}, and \texttt{data\_source}.
The themes and solution turns from the source conversation are
deliberately unused for this toy task.

\paragraph{System prompt.}
The training-time system prompt mandates a numeric estimate (refusal is
explicitly forbidden), tells the agent which lc0 tools are available,
and requires the reply to end with the literal line
\verb|The popularity is <int> and the ELO is <int>|. Without the explicit
``refusing is forbidden \ldots\ guess if you must'' instruction, Qwen3-4B
defaults to ``I cannot directly determine \ldots'' and never emits a tool
call --- format-pass rate collapses to 0\%.

\paragraph{Reward.}
Format-gated multiplicative regression:
\begin{align*}
\hat p, \hat e &= \mathrm{parse}(\text{rollout completion}),\\
f               &= \mathbb{1}\bigl[\hat p \neq \texttt{None} \,\wedge\, \hat e \neq \texttt{None}\bigr],\\
R_p             &= \max\!\bigl(0,\, 1 - |\hat p - p^\star| / 50\bigr),\\
R_e             &= \max\!\bigl(0,\, 1 - |\hat e - e^\star| / 400\bigr),\\
R               &= f \cdot \tfrac{1}{2}\bigl(R_p + R_e\bigr).
\end{align*}
Scales of 50 (popularity) and 400 (ELO) give linear partial credit out
to roughly half the natural range of each target before clamping. The
multiplicative format gate ($f$) ensures that an answer-shaped scribble
that misses the canonical line scores zero --- this prevents the
``answer-shaped output'' reward hack documented in \texttt{rl\_recipes.md}
\S2. Read-only tool calls earn no separate reward, in line with the
Iterative Reward Calibration finding that positive per-call bonuses
drive over-invocation.

The parser at \texttt{rewards/parser.py} matches case-insensitively, allows
the article ``the'' before \texttt{ELO} to be omitted (a common drift),
and rejects sentence-break separation (``\textsl{popularity is X. The ELO
is Y.}''), since the system prompt explicitly mandates one line. If the
rollout contains multiple matches the \emph{last} one is taken, on the
assumption that the model often shows scratch work before its final
answer.

\paragraph{Rollout smoke test.}
Before running RL we verify the harness end-to-end on 20 held-out
puzzles using the exact prompt and reward that training will see. The
agent invokes lc0 tools through \texttt{agents/chess\_analyst.py} and the
final completion is scored with \texttt{rewards.puzzle\_reward.compute\_score}.

\begin{table}[h]
\centering
\small
\begin{tabular}{lr}
\toprule
\textbf{Metric} & \textbf{Value (20 puzzles, $T{=}0$)} \\
\midrule
Format-pass rate          & 100.0\% \\
Mean reward               & 0.432 \\
Reward standard deviation & 0.220 \\
Mean tool calls per run   & 2.0 \\
Mean LLM rounds per run   & 3.0 \\
Wall time per puzzle      & 1.9\,s \\
\bottomrule
\end{tabular}
\caption{Rollout smoke-test stats. The non-zero reward variance is the
quantity that matters for GRPO: groups of rollouts that share a single
prompt but yield different rewards generate non-zero advantages and
therefore non-zero gradient. The greedy decoding bias toward a fixed
answer (the model frequently outputs $75/2200$) does not collapse the
gradient because the ground truth varies across the group.}
\label{tab:smoke-stats}
\end{table}

%------------------------------------------------------------------------------
\subsection{Prompt Templates}
\label{app:details:prompts}

% TODO: list training-time prompts (GRPO reward prompt, projector pretraining format,
%       InvokeAgent trigger format, multi-turn template).
% Cross-reference with pages/appendix/prompts.tex for extended listings.
