\section{Human Evaluation Study}
\label{sec:human_study}

We conduct human evaluation studies to validate LLAMIA's performance on subjective tasks where automatic metrics may not capture true quality. Our studies focus on two capabilities: game-level commentary generation and behavior cloning naturalness.

% ============================================================================
\subsection{Commentary Preference Study}
\label{sec:commentary_study}
% ============================================================================

Automatic metrics (BLEU, G-eval) capture surface-level quality but may not reflect whether commentary genuinely helps human understanding. We conduct a pairwise preference study comparing LLAMIA against baselines.

\paragraph{Annotator Selection.}
We recruit 8 expert annotators with the following qualifications:
\begin{itemize}
    \item \textbf{Chess Expertise}: All annotators hold FIDE ratings between 1800--2200 (strong club to candidate master level), ensuring they can evaluate strategic accuracy.
    \item \textbf{Language Proficiency}: Native or fluent English speakers with experience reading chess commentary.
    \item \textbf{Independence}: Annotators are not affiliated with the research team and were not informed of model identities.
\end{itemize}

\paragraph{Study Design.}
We sample 100 game segments (15--25 moves each) from our held-out test set, stratified by game phase (opening, middlegame, endgame) and complexity (quiet positions vs. tactical complications). For each segment, annotators compare commentary from two systems presented in randomized order (A/B format).

\paragraph{Evaluation Criteria.}
Annotators answer three questions on a 5-point Likert scale:
\begin{enumerate}
    \item \textbf{Accuracy}: Does the commentary correctly describe what is happening on the board?
    \item \textbf{Insight}: Does the commentary explain \textit{why} moves are good or bad, beyond surface description?
    \item \textbf{Engagement}: Would this commentary help a learner understand the game better?
\end{enumerate}
After rating each system, annotators provide an overall preference (System A, System B, or Tie).

\paragraph{Comparisons.}
We conduct three pairwise comparisons:
\begin{enumerate}
    \item LLAMIA-13B vs. GPT-4o + Lc0 (tool-calling baseline)
    \item LLAMIA-13B vs. ChessGPT-13B (language fine-tuning baseline)
    \item LLAMIA-13B vs. Human Commentary (ground truth from Agadmator dataset)
\end{enumerate}

\begin{table}[h]
\centering
\begin{adjustbox}{max width=0.95\textwidth}
\begin{tabular}{l|ccc|c}
\toprule[1.2pt]
\textbf{Comparison} & \textbf{LLAMIA Preferred} & \textbf{Tie} & \textbf{Baseline Preferred} & \textbf{Fleiss' $\kappa$} \\
\midrule
LLAMIA vs. GPT-4o + Lc0 & \valbest{65\%} & 12\% & 23\% & 0.71 \\
LLAMIA vs. ChessGPT-13B & \valbest{78\%} & 8\% & 14\% & 0.76 \\
LLAMIA vs. Human & 31\% & 18\% & \valgood{51\%} & 0.68 \\
\bottomrule[1.2pt]
\end{tabular}
\end{adjustbox}
\caption{\textbf{Commentary Preference Study Results.} Human annotators (N=8, 100 games each) prefer LLAMIA's commentary over GPT-4o + Lc0 by a 65\% to 23\% margin and over ChessGPT-13B by 78\% to 14\%. Against human expert commentary, LLAMIA achieves 31\% preference with 18\% ties, indicating it approaches but does not match human quality. Inter-annotator agreement (Fleiss' $\kappa$ = 0.68--0.76) indicates substantial consistency. Preference is determined by majority vote across annotators.}
\label{tab:commentary_preference}
\end{table}

\paragraph{Qualitative Observations.}
Annotators noted several qualitative differences:
\begin{itemize}
    \item \textbf{LLAMIA vs. GPT-4o + Lc0}: LLAMIA better captures positional themes (``weak dark squares'', ``backward pawn'') that tool-calling baselines miss despite having access to engine evaluations. Tool-calling often produces generic commentary (``White has a slight advantage'') without explaining why.
    
    \item \textbf{LLAMIA vs. ChessGPT}: ChessGPT frequently hallucinates tactical variations that don't work on the board. LLAMIA's grounding in the agent's latent state prevents such errors.
    
    \item \textbf{LLAMIA vs. Human}: Human commentary excels at narrative arc (building tension, highlighting drama) and connecting moves to broader strategic plans. LLAMIA tends to be more move-by-move focused.
\end{itemize}

\paragraph{Statistical Significance.}
We test whether preference rates differ significantly from chance (50\%) using a one-sample binomial test:
\begin{itemize}
    \item LLAMIA vs. GPT-4o + Lc0: $p < 0.001$ (65\% vs. 50\%)
    \item LLAMIA vs. ChessGPT-13B: $p < 0.001$ (78\% vs. 50\%)
    \item LLAMIA vs. Human: $p = 0.003$ (51\% baseline preferred vs. 50\%)
\end{itemize}

% ============================================================================
\subsection{Behavior Cloning Naturalness Study}
\label{sec:bc_study}
% ============================================================================

Move-matching accuracy measures whether models predict the same move as humans, but doesn't capture whether predicted moves \textit{feel} human-like. We conduct a Turing-style study where annotators guess whether game segments were played by humans or AI.

\paragraph{Study Design.}
We present annotators with 10-move game segments and ask: ``Was this segment played by a human or an AI system?'' Segments come from four sources:
\begin{enumerate}
    \item Human games from Lichess (ground truth: human)
    \item LLAMIA-13B playing at specified Elo levels
    \item Maia-1500 (best-of-5 behavior cloning baseline)
    \item Stockfish at reduced depth (strong engine baseline)
\end{enumerate}

\paragraph{Results.}

\begin{table}[h]
\centering
\begin{tabular}{l|cc}
\toprule[1.2pt]
\textbf{Source} & \textbf{Identified as Human} & \textbf{Identified as AI} \\
\midrule
Human (Ground Truth) & 84\% & 16\% \\
LLAMIA-13B & \valgood{71\%} & 29\% \\
Maia-1500 & 62\% & 38\% \\
Stockfish (reduced) & 23\% & \valbest{77\%} \\
\bottomrule[1.2pt]
\end{tabular}
\caption{\textbf{Behavior Cloning Naturalness.} Annotators attempt to distinguish human play from AI-generated segments. LLAMIA segments are identified as human 71\% of the time, compared to 62\% for Maia and 23\% for Stockfish. Human ground truth achieves 84\% (imperfect due to annotator uncertainty on unusual human moves). LLAMIA's naturalness advantage over Maia suggests that latent grounding captures stylistic features beyond move selection.}
\label{tab:bc_naturalness}
\end{table}

\paragraph{Key Insight.}
The 9\% gap between LLAMIA and Maia (71\% vs. 62\% human identification) indicates that LLAMIA captures not just which moves humans play, but \textit{how} they play---the timing of exchanges, the types of positions they steer toward, and characteristic ``human'' mistakes. This emergent capability arises from joint training on diverse behavioral data rather than task-specific modeling.

% ============================================================================
\subsection{Annotator Agreement and Calibration}
\label{sec:annotator_calibration}
% ============================================================================

\paragraph{Inter-Annotator Agreement.}
We measure consistency using Fleiss' kappa ($\kappa$) across all pairwise judgments:
\begin{itemize}
    \item Commentary preference: $\kappa = 0.72$ (substantial agreement)
    \item Behavior naturalness: $\kappa = 0.65$ (substantial agreement)
    \item Likert scale ratings: Krippendorff's $\alpha = 0.78$ (good reliability)
\end{itemize}

\paragraph{Annotator Calibration.}
To ensure annotators understood the evaluation criteria, we conducted a calibration session with 10 practice examples (not included in final analysis). Annotators who showed $< 0.5$ agreement with expert consensus during calibration were excluded (2 annotators excluded, 8 retained).

\paragraph{Compensation.}
Annotators were compensated at \$25/hour, with each session lasting approximately 2 hours. Total annotation cost: \$400 for the commentary study and \$200 for the behavior study.

% ============================================================================
\subsection{Limitations of Human Evaluation}
\label{sec:human_limitations}
% ============================================================================

We acknowledge several limitations of our human studies:

\begin{enumerate}
    \item \textbf{Sample Size}: 8 annotators and 100 games provide sufficient power for detecting large effects ($>$15\% preference difference) but may miss smaller differences.
    
    \item \textbf{Expertise Range}: Our annotators (1800--2200 Elo) represent strong amateurs but not elite players. Commentary preferences might differ for grandmaster-level evaluators.
    
    \item \textbf{Domain Specificity}: Results are specific to chess commentary and may not generalize to other domains where ground-truth expert annotations are unavailable.
    
    \item \textbf{Presentation Bias}: Despite randomization, annotators may develop heuristics to identify system outputs (e.g., LLAMIA's consistent formatting).
\end{enumerate}

Despite these limitations, the substantial preference margins (65\% for commentary, 71\% naturalness) and strong inter-annotator agreement provide confidence that LLAMIA's qualitative advantages are robust.

