\section{Paper TODO}
\label{sec:todo}

Working TODO for the camera-ready. Items are tagged \textbf{[P0]}
(camera-ready blocker), \textbf{[P1]} (paper materially weaker without
it), \textbf{[P2]} (appendix-only / nice-to-have). Each item names the
positioning or argument it serves so descoping decisions stay
principled. Reference briefs (positioning, metric definitions, baseline
spec, headline figure spec) are at the end of this document; writing
items point to them.

% =============================================================================
\subsection{Experiments}
\label{sec:todo_experiments}
% =============================================================================

The paper makes four claims. Each claim has a single quantitative
anchor experiment in the main paper; setups, breadth checks,
sensitivity sweeps, and qualitative galleries live in the appendix and
are referenced from discussion. Every main-paper quantitative
experiment carries one matched-recipe ablation in the appendix; if it
does not, it does not belong in the main paper.

\textbf{Claims.}
\textbf{C1} -- LLAMIA-Bench provides comprehensive ground-truth-anchored
evaluation of LM--specialist collaboration across seven facets, in a
single environment;
\textbf{C2} (corollary) -- the verbalisation gap survives matched
backbone, matched compute, and matched RL recipe -- it is an
\emph{interface} property, not a scale or recipe artefact;
\textbf{C3} -- dynamic re-encoding is the load-bearing mechanism of
the latent-internalisation interface;
\textbf{C4} -- the latent space is interpretable and instruction-steerable
in feature space (instruction reproduces what fine-tuning does);
\textbf{C5} -- the integration paradigm is not chess-specific.

\subsubsection*{Claim-anchor experiments [P0]}

\begin{itemize}[leftmargin=*]
    \item \textbf{[C1] Full LLAMIA-Bench evaluation.} All baselines
    (Groups A--F, brief~\ref{brief:baselines}) over the seven-facet
    suite; one task per facet in the main-paper table
    (\textsc{Move-Class}, \textsc{Eval-Rat}, \textsc{Annot-Move},
    \textsc{Comm-Game}, \textsc{BC-Maia}, \textsc{ICP-Win},
    \textsc{ICP-Adversarial}). VG / DCS / RY scalars
    (brief~\ref{brief:metrics}) summarise the integration-interface
    corollary on this same table -- no separate experiment.
    \emph{Purpose:} the single source of truth for every quantitative
    claim downstream; the headline radar visualises this row-set.

    \item \textbf{[C2] Three matched RL runs:
    LLAMIA-text / LLAMIA-tool / LLAMIA-full.}
    Identical backbone, reward, rollout count, RL recipe; only the
    integration mode differs. Reported as one row block in the C1
    table. See brief~\ref{brief:methodology}.
    \emph{Purpose:} settles the interface-vs-scale-vs-recipe confound
    that frontier-system comparisons (Groups A/B) cannot.

    \item \textbf{[C3] Latent vs.\ symbolic-rich, matched information
    content.} Replace $\vz_s$ with a textual representation of equal
    information content (top-$k$ moves with full eval, WDL, PVs to
    depth 20, positional features); train at matched FLOPs.
    \emph{Purpose:} the cleanest possible ablation of representation
    \emph{form} at matched information \emph{content}. If symbolic-rich
    text closes the gap, C3 fails. The within-system
    \texttt{LLAMIA-static} (no \texttt{<InvokeAgent>}) row of the C1
    table is the matched paired ablation reported in the same place.

    \item \textbf{[C4] Causal steering via a single latent
    direction.} Modeled on the single-method discipline of
    \cite{Leela-lookahead-8zg9sO4ttV}: one method, one curated
    position set, one headline number.
    \emph{Position set:} $N$ positions where adding an ICP
    instruction $I$ flips LLAMIA-full's move from $a$ to $a'$ (the
    instructed move).
    \emph{Direction:} a single direction $\vd_I \in \mathbb{R}^{1024}$
    in $\vz_s$-space identified per instruction theme (the SAE is the
    \emph{instrument} for finding $\vd_I$, not the headline).
    \emph{Intervention:} strip the language instruction from the
    prompt -- LLAMIA reverts to $a$; then \emph{add} $\vd_I$ to
    $\vz_s$ at fixed magnitude.
    \emph{Headline metric:} fraction of positions on which adding
    $\vd_I$ alone (no language instruction in context) recovers the
    instructed move $a'$; reported against a random-direction null
    and the full-instruction ceiling. Single number per ICP theme,
    aggregated across themes.
    \emph{Single main figure:} recovery rate per theme with
    random-direction null bars and full-instruction ceiling line;
    one qualitative position panel inset.
    \emph{Purpose:} the cleanest causal statement of C4 -- a
    \emph{single} latent direction reproduces the policy effect of
    the language instruction. ``Instruction reproduces what
    fine-tuning does'' becomes one experiment, one method, one
    number, not a five-method bundle.

    \item \textbf{[C5] Cross-domain extension.} \emph{Poker} on
    OpenSpiel (Texas Hold'em or Leduc): hidden information and
    bluffing add what chess cannot show; one main-paper figure
    comparing LLAMIA-poker, the same backbone with verbalised tool
    access to a poker specialist, and frontier baselines on a single
    opponent-modelling task. Vision and persuasion extension results
    move to \S\ref{sec:cross_domain} / appendix as already-existing
    evidence.
    \emph{Purpose:} demonstrates the paradigm is not chess-specific;
    one figure, one paragraph.

    \item \textbf{[C2 visual] Three-method GRPO training curves.}
    Reward and \textsc{ICP-Win} win-rate vs.\ training step for the
    three matched RL runs.
    \emph{Purpose:} the methodology figure; replaces the legacy
    \texttt{kt}/\texttt{comm}/\texttt{icp} subfigure trio. Single
    figure; setup details to appendix.

    \item \textbf{[C5 secondary] LLM size $\times$ agent strength
    heatmap.} LLM $\in \{1, 3, 7, 13, 32\text{B}\}$ $\times$ agent
    $\in \{$Maia-1500, Lc0-T72, T80, BT4, Stockfish~17$\}$; one panel
    per facet (perception / sustained planning / instruction /
    explanation).
    \emph{Purpose:} the ``insight that travels'' across LM and agent
    scale; argues the paradigm benefits both sides of the integration
    monotonically. One figure; numerical detail in appendix.
\end{itemize}

\subsubsection*{Supporting [P1, brief mention in main, full in appendix]}

\begin{itemize}[leftmargin=*]
    \item \textbf{Group~F matched-recipe ablations.}
    \texttt{LLAMIA-static}, \texttt{Proj-Frozen}, \texttt{No-Proj}
    rows. One sentence in main; full table in
    \texttt{appendix/ablations.tex}.

    \item \textbf{Group~G Lc0 architecture cross-check.}
    BT4 / T60 / smaller Lc0 nets. One sentence in discussion
    (``the integration story holds across Lc0 architectures, see
    Tab.~X''); single appendix table; no main-text figure.

    \item \textbf{Maia-equivalence at policy level.} For
    $X \in \{1100,1300,1500,1700,1900\}$: top-1 match, KL to
    actual-human-Elo-X moves, achieved-Elo. Maia-1900 row already
    appears in the C1 table; the full per-Elo curve is appendix.

    \item \textbf{\textsc{ICP} turn-by-turn win-rate curve.}
    Sustained-interaction signature: gap widens with horizon. One
    appendix figure; one sentence in main.

    % Lichess deployment removed: LLM inference latency makes real-time play infeasible.

    \item \textbf{Stylometry blind-tasting.} Top-1/2/3 with held-out
    real GMs. One number in main (top-1 with CI); confusion matrix and
    Bradley--Terry baseline (\cite{CvjXlsBLCX-strength-est}) appendix.

    \item \textbf{Verbalised-tool failure taxonomy} on Groups A/B/C
    via the F1--F6 codebook (brief~\ref{brief:failure}).
    Stacked-bar figure in main; full coding rubric and
    inter-annotator $\kappa$ appendix.

    \item \textbf{Pre-cutoff vs.\ post-cutoff contamination test.}
    Per-baseline cutoff date; $z$/Fisher tests on the gap. One row
    in the rigor audit (\S\ref{sec:todo_infra}); appendix detail.

    \item \textbf{Group~D latent-CoT baselines.} Coconut, latentmas,
    2025--2026 implicit-reasoning baselines. One row block in the C1
    table; setup and prompt details appendix.

    \item \textbf{Human preference study.} Rated chess players
    ($\geq$1700 Elo, $n \geq 60$) see side-by-side outputs from
    LLAMIA and the strongest verbalized baseline on matched positions
    (commentary + move rationale). Raters choose which explanation
    they would trust to improve their own play, blind to system
    identity. Preference rate (with 95\% CI) is the single reported
    number; one figure in main. Simultaneously validates subjective
    quality and grounds our judge-based metrics: rater agreement with
    GPT-4o scores on the same items ($\kappa$) is reported in the
    caption.
    \emph{Purpose:} one study that both demonstrates human-perceived
    quality of LLAMIA over verbalized tool-calling \emph{and}
    validates the LLM judge, without needing separate experiments.
\end{itemize}

\subsubsection*{Appendix-only / qualitative galleries [P2]}

Items below are referenced \emph{en bloc} from discussion (e.g.,
``further sensitivity sweeps and qualitative examples in
Appendix~\ref{sec:ablations}''); none deserve their own paragraph in
the main text.

\begin{itemize}[leftmargin=*]
    \item \textbf{Sensitivity sweeps.} Projector capacity (linear /
    3-MLP / 6-MLP / Q-Former); loss-weighting; test-time-compute
    Pareto (\#\texttt{<InvokeAgent>} per game $\times$ task score);
    data-axis scaling (Stage-1 $\in\{1,5,25,50\}$M; Stage-2
    $\in\{50\text{k},250\text{k},1\text{M}\}$); compute-allocation
    Pareto (LM-side vs.\ agent-side search budget); LoRA vs.\ full
    FT; top-$p$/temperature robustness; reasoning-budget
    $\times$ task difficulty.
    \item \textbf{Latent-form controls.} $\vz_s$ vs.\ $\vh_s$
    (raw vs.\ projected, token-padded); random / zeroed / corrupted
    / weak-engine latents; future-paradigm latent ablation
    (S-AA / S-ASA / S-AVAV / S-AVSAV).
    \item \textbf{Diagnostic decompositions.} Phase-of-game
    (opening/middle/end); $k$-step rollout BC at
    $k\in\{1,3,5,8\}$; multi-step commentary fidelity vs.\ horizon;
    audience conditioning (1200 vs.\ 2200 prompts);
    verbosity-binned commentary scores; move-level vs.\ game-level
    commentary delta on the same model; three-stage CoT-error
    decomposition (\cite{yRxX01oRIi-cot-hurts}) on verbalised-tool
    baselines.
    \item \textbf{Difficulty (Spearman-only).} Spearman-$\rho$
    against Lichess puzzle ratings on \textsc{Diff-1pos} /
    \textsc{Diff-Rat}. The rater-judged \emph{rationale-correctness}
    branch is dropped; the rationale contribution is already covered
    by \textsc{Annot-Move} / \textsc{Comm-Move} under Explanation.
    \item \textbf{Hard-puzzle subset.} Top-decile Lichess
    ($\sim$3000+) and an evergreen ``unseen-after-$T$'' subset.
    Anti-saturation and contamination control. One appendix table.
    \item \textbf{Round-robin Elo table.} LLAMIA-instructed-as-X /
    Maia-X / depth-matched Lc0 cells; iterative Elo per
    \cite{CvjXlsBLCX-strength-est}. Move-match accuracy is already
    the primary public-benchmark anchor in main; this lives in
    appendix as the strength-conversion check.
    \item \textbf{Counterfactual-coverage example.} One annotated
    game; subset of gold commentary discussing counterfactual lines;
    LLAMIA vs.\ verbalised-tool baseline coverage. Qualitative
    illustration of why dynamic re-encoding helps explanation.
    \item \textbf{Strategy Sankey for ICP themes.} Qualitative
    storytelling; one appendix figure.
    \item \textbf{Robotics / Hanabi pilot.} Stretch extension
    referenced as ``ongoing.''
\end{itemize}

\subsubsection*{Removed (with brief reasons)}

\begin{itemize}[leftmargin=*]
    \item \textbf{Cross-Elo extrapolation} (train on $\{1300,1700\}$,
    test on $\{1100,1500,1900\}$). The objective requires too much
    chess-specific setup to land cleanly with a non-chess reader; the
    representation-form claim is already cleanly delivered by the
    [C3] symbolic-rich-vs-latent ablation.
    \item \textbf{Counter-intuitiveness signal} via shallow-vs-deep
    Lc0 disagreement. Search-depth disagreement is a noisy proxy for
    difficulty and the argument requires a multi-paragraph chess
    aside; the Lichess-rating Spearman in the appendix difficulty
    sweep already does the calibration job.
    \item \textbf{Difficulty + rationale joint scoring (rationale
    branch).} The rater-judged rationale-correctness score is
    expensive, subjective, and redundant with \textsc{Annot-Move} /
    \textsc{Comm-Move}; only the Spearman branch survives, in
    appendix.
    \item \textbf{AIME-style OOD micro-experiment.} Out of scope
    given the [C5] poker extension already delivers the
    cross-domain claim; AIME adds a non-chess, non-game reader load
    for marginal evidence.
\end{itemize>

% =============================================================================
\subsection{Analyses}
\label{sec:todo_analyses}
% =============================================================================

Two analyses live in the main paper -- the bench scalars (VG / DCS /
RY, brief~\ref{brief:metrics}) reported alongside the [C1] table, and
the single causal-steering protocol that \emph{is} the [C4] anchor
experiment. Modeled on \cite{Leela-lookahead-8zg9sO4ttV}'s
single-method discipline: one method, one curated position set, one
headline number; convergent-method checks are appendix support, not
co-headlines. Everything else is appendix.

\subsubsection*{Main-paper analyses [P0]}

\begin{itemize}[leftmargin=*]
    \item \textbf{Bench scalars (VG / DCS / RY).} Computed directly
    from the [C1] table; reported in the same table caption with
    bootstrap CIs. No separate experimental run.

    \item \textbf{Causal-steering protocol (the [C4] method, in
    full).} The single method behind the C4 anchor experiment;
    everything else interpretability-related is appendix support.
    \emph{Position set construction:} $N\!\geq\!200$ positions per
    instruction theme where ICP-instructed LLAMIA picks
    $a' \neq a$ and base LLAMIA picks $a$; balanced across phases
    and material; held out from training.
    \emph{Direction identification:} train an SAE on $\vz_s$ over
    the Stage-1 corpus (top-$k$ or JumpReLU; dictionary
    $\in\{8\text{k},32\text{k}\}$, $L_0\!\approx\!32$, ghost-grad;
    hyperparameters chosen for reconstruction loss only, not
    downstream metric, to avoid circularity); for each instruction
    $I$, fit a sparse linear classifier over SAE activations to
    predict the binary instruction-vs-base label on the position
    set; the largest-magnitude coefficient defines $\vd_I$ (a
    \emph{single} SAE direction). The SAE is the dictionary, not
    the result.
    \emph{Intervention and metric:} ablate the language instruction
    from the prompt; add $\vd_I$ to $\vz_s$ at the magnitude that
    matches the average $\|\Delta\vz_s\|$ between instructed and
    base on a held-out tuning split; report (i) instructed-move
    recovery rate, (ii) random-direction null at the same magnitude
    ($\geq 50$ random unit-norm directions, percentile reported),
    (iii) full-instruction ceiling, (iv) \textsc{ICP-Win} win-rate
    recovery on a held-out game set when the language instruction
    is replaced by the steering vector throughout play.
    \emph{Pre-registered success criterion:} recovery
    $\geq 0.5\!\times$ ceiling and $> 95$th percentile of
    random-direction null on $\geq 4$ of 5 instruction themes.

    \item \textbf{Train/test data-leakage audit.} Per benchmark: FEN
    overlap, instruction overlap, Lichess game-ID overlap, Stage-1
    self-play date cutoff. Single audit table; mandatory.
\end{itemize}

\subsubsection*{Appendix-only analyses [P1/P2]}

Referenced in main text only as ``further analyses in
Appendix~\ref{sec:ablations}.'' The first three items are the
convergent-method checks dropped from the main paper to keep [C4]
single-method; they appear as one appendix subsection (``Convergent
evidence for the steering direction'') answering ``is the same
$\vd_I$ visible from independent vantage points.''

\begin{itemize}[leftmargin=*]
    \item \textbf{[Convergent] Linear probes for chess concepts on
    $\vz_s$ and $\vh_s$.} \cite{BLOIB8CwBI-soundness}-style
    decodability sanity; one table.
    \item \textbf{[Convergent] Activation patching} between (clean,
    corrupted) FEN pairs where Lc0's top move flips but a weak
    engine's does not (\cite{Leela-lookahead-8zg9sO4ttV}-style);
    locates which $\vz_s$ token positions carry the steering
    signal.
    \item \textbf{[Convergent] $\pi$-vs-$\pi'$ feature-delta.}
    SAE-decompose $\Delta\vz_s$ between base, ICP, and Maia-X
    instruction conditions on a fixed position; show sparse
    $\Delta f$ overlapping with the chosen $\vd_I$.
    \item \textbf{Future-square probes} on $\vz_s$ for
    principal-variation squares at 1, 3, 5 moves ahead
    (\cite{Leela-lookahead-8zg9sO4ttV} \S 3.3 adaptation; mate-in-$N$
    vs.\ quiet split).
    \item \textbf{Per-token attribution.} Ablate each of $K{=}16$
    tokens; plot log-odds reduction; group tokens by ablation
    effect.
    \item \textbf{Adversarial soundness in commentary.}
    Factual-board-error count per 1000 commentary tokens with and
    without dynamic re-encoding; quantifies the F5 elimination claim.
    \item \textbf{First-$N$ non-cherry-picked SAE feature catalog.}
    First 40 features by firing frequency annotated; remainder behind
    anonymized URL (\cite{chessformer-2ltBRzEHyd}-style).
    \item \textbf{Human ecological-validity check on SAE features.}
    Two annotated games to chess-experienced raters; identify which
    followed an instruction.
    \item \textbf{LLAMIA-side failure-mode analysis.}
    LLM-side / agent-side / interface mis-conditioning; confusion
    matrix; balances the verbalised-tool taxonomy.
    \item \textbf{Communication Efficiency (CE) per task.}
    Score / (LLM tokens + InvokeAgent calls + tool-call tokens);
    pairs with the test-time-compute Pareto.
    \item \textbf{Verbosity-conditional commentary score.}
    Re-express commentary scores conditional on token count; defends
    against ``wins by writing more.''
\end{itemize>

% =============================================================================
\subsection{Writing and figures}
\label{sec:todo_writing}
% =============================================================================

\subsubsection*{P0 -- writing}

\begin{itemize}[leftmargin=*]
    \item \textbf{Rewrite the LLAMIA-Bench section per
    brief~\ref{brief:bench}.} Translator metaphor as the opening; one
    organizing axis (interaction depth); the falsifiable, counter-
    intuitive prediction; the three custom metrics; the failure
    taxonomy; public benchmarks separate; real-world relevance;
    explicit ``what this is not''; acknowledged limitations; novelty
    paragraph; minimum-viable-bench fallback.
    \emph{Purpose:} the bench section currently lacks soul; the brief
    is the spec.

    \item \textbf{Rewrite the methodology section as
    SALT(SFT) + SALT(RL) per brief~\ref{brief:methodology}.}
    SFT becomes the warm-start; RL is the headline. Reward
    decomposition: outcome, format, collaboration-shape.
    \emph{Purpose:} 2026-vintage methodology; SFT-only is no longer
    publishable as the headline.

    \item \textbf{Rewrite the baselines section per
    brief~\ref{brief:baselines}.} Seven groups (A--G); explicit
    access-mode and recipe tags; protocol box for tool calling.
    \emph{Purpose:} replaces the 2024-vintage baseline grid;
    establishes Group~C as the strawman the thesis attacks.

    \item \textbf{Notation cleanup.} Single canonical set
    $\{\vs_t, \va_t, \vz_t\}$; fix the loss-formulation language
    (``conditioned on tokens it is also tasked with predicting''
    is the most damaging prior-reviewer complaint).
    \emph{Purpose:} reviewer ZW8t / HUaR; eliminates the most concrete
    technical objection from the prior round.

    \item \textbf{POMDP framing.} Either a precise decision-theoretic
    derivation or downgrade to ``inspired by.''
    \emph{Purpose:} reviewer-defensive on theoretical framing.

    \item \textbf{Tool-call protocol box (appendix).} Pseudocode +
    prompts + retry policy + decoding settings.
    \emph{Purpose:} reviewer HUaR demand; reproducibility.

    \item \textbf{Latent-CoT related work.} Coconut, latentmas,
    2025--2026 implicit-reasoning literature.
    \emph{Purpose:} closes the most obvious related-work gap.

    \item \textbf{Robotics / autonomous-driving related work.}
    RT-2, PaLM-E, Gato, Octo.
    \emph{Purpose:} reviewer HUaR demand; broadens the audience.

    \item \textbf{Distinct-from-Karvonen paragraph in interpretability
    section.} We operate on the LLM-side projection of an external
    policy, not the policy's own latents.
    \emph{Purpose:} mandatory differentiation from the chess-SAE prior
    work that got rejected.

    \item \textbf{Threats-to-validity paragraph.}
    \emph{Purpose:} standard rigor; pre-empts methodological doubt.

    \item \textbf{``What this benchmark is not'' paragraph.}
    Already in brief~\ref{brief:bench}; ensure it lands in the bench
    section verbatim.

    \item \textbf{Banned-phrase sweep.} ``scaling laws,'' ``compounding
    error'' (use ``error amplification''), ``simultaneously masters,''
    ``counterparts.''
    \emph{Purpose:} pre-empts framing nitpicks that have sunk similar
    papers (\cite{DiffuSearch-A9y3LFX4ds} reject).

    \item \textbf{Resolve all \texttt{\textbackslash somesh\{\dots\}}
    margin notes.}
    \emph{Purpose:} draft hygiene.

    \item \textbf{Cross-link every Introduction claim to a backing
    table or figure.}
    \emph{Purpose:} TMLR-style claims-match-evidence audit.
\end{itemize}

\subsubsection*{Main-paper figures and tables [P0]}

Hard cap: every claim gets one figure or one table, not both.

\begin{itemize}[leftmargin=*]
    \item \textbf{[C1] Headline radar.} Seven spokes grouped into
    seven facet-arcs (perception, calibration, explanation, sustained
    planning, style/identity, instruction-conditioned adaptation,
    adversarial robustness); three systems overlaid
    (\texttt{LLAMIA-full}, the strongest Group~C verbalised-tool-RL
    baseline, the strongest Group~A frontier reasoning model);
    per-spoke specialist ceilings as separate dots; depth tag
    (D1/D2/D3) as a secondary spoke annotation. See
    brief~\ref{brief:radar}.

    \item \textbf{[C1] LLAMIA-Bench results table.} One row per
    system (Groups A--F), one column per facet plus VG / DCS / RY
    summary columns; the body the radar summarises.

    \item \textbf{[C1] Public-benchmark anchor table.} One row per
    public benchmark (Maia-1900 BC, ChessBench Elo, Lichess solve
    rate, ChessQA, HPDv2). Sits next to the radar so LLAMIA cannot be
    dismissed as a niche-bench winner.

    \item \textbf{[C2] Three-method GRPO training curves.} Reward
    and \textsc{ICP-Win} vs.\ training step. The methodology figure.

    \item \textbf{[C2] Verbalised-tool failure-mode stacked bar.}
    F1--F6 across integration modes per facet category;
    visualises ``F1--F5 are exclusively verbalisation failures.''

    \item \textbf{[C4] Causal-steering recovery figure.} Bar chart
    of instructed-move recovery rate per ICP theme under three
    conditions: random-direction null (with 95\% percentile band),
    single-direction $\vd_I$ steering (no language instruction),
    and full-language-instruction ceiling. One inset position panel
    showing the move flip under steering. The interpretability
    headline -- one method, one number per theme.

    \item \textbf{[C5] LLM size $\times$ agent strength heatmap.}
    Four facet panels.

    \item \textbf{[C5] Cross-domain extension figure.} One panel
    (poker opponent-modelling task); supporting vision/persuasion
    table referenced from \S\ref{sec:cross_domain}.
\end{itemize}

\subsubsection*{Appendix figures and tables [P1/P2]}

\begin{itemize}[leftmargin=*]
    \item Per-Elo move-match curves with binomial CIs (Maia
    equivalence).
    \item \textsc{ICP} turn-by-turn win-rate curve (sustained-
    interaction signature).
    \item Round-robin Elo table (LLAMIA-as-X / Maia-X /
    depth-matched Lc0).
    \item Replay graph for game-level commentary (engine-eval
    trajectory + commentary salience + annotation-accuracy overlay).
    \item Strategy Sankey for ICP themes (qualitative).
    \item Per-move strength curves (BC sustained-quality).
    \item LLAMIA-side failure-mode confusion matrix.
    \item CC-AUC heatmap for SAE features.
    \item Verbosity-binned commentary comparison.
    \item $k$-step rollout BC degradation curve.
    \item Feature-frequency / dead-feature histogram.
    \item Cost-per-game (\$ / tokens) panel.
\end{itemize}

\subsubsection*{P1 -- writing housekeeping}

\begin{itemize}[leftmargin=*]
    \item \textbf{Color-coding legend / SOTA(SFT) named in tables.}
    Reviewer LEvg.
    \item \textbf{Limitations expansion.} \cite{lnIlH0hfek-mom}
    reject lesson; the prior limitations section is too thin.
    \item \textbf{Concrete bench description in abstract and intro
    contributions.} Name the seven facets and one task per facet so
    claims and evidence match (\cite{TIGQIem1na-comma} reviewer Pdtk
    lesson).
\end{itemize}

% =============================================================================
\subsection{Infrastructure and rigor}
\label{sec:todo_infra}
% =============================================================================

\subsubsection*{P0}

\begin{itemize}[leftmargin=*]
    \item \textbf{Populate \texttt{appendix/dataset.tex}.} Formal task
    definitions for every D1/D2/D3 task: prompt template, gold
    rationale construction, scoring rubric, train/val/test split,
    contamination control, sample size.
    \emph{Purpose:} the bench currently exists only as names;
    reviewers will look here.

    \item \textbf{Pre-register failure-mode classifier.}
    Prompts + 100-conversation human-rater calibration set; OSF or
    AsPredicted.
    \emph{Purpose:} LLM-judge bias is the standard reviewer attack on
    bench papers; pre-registration neutralizes it.

    \item \textbf{Pre-register human-study analysis plan.}
    \emph{Purpose:} same neutralization for the human-eval claims.

    \item \textbf{IRB exemption.} 45 CFR 46.104(d)(2)(ii) for the
    Lichess deployment and the human-rater study.

    \item \textbf{95\% bootstrap CIs ($\geq 500$ resamples) on every
    headline metric.}

    \item \textbf{Binomial / $z$ / Fisher tests for human-eval.}

    \item \textbf{Inter-annotator $\kappa$ for every qualitative
    coding} ($\geq 0.7$ target).

    \item \textbf{Matched-pair LLM comparisons.} Same family, adjacent
    release dates only.
    \emph{Purpose:} prevents apples-to-oranges accusations.

    \item \textbf{Player- (or game-) disjoint train/test splits.}
    \emph{Purpose:} eliminates the most damaging contamination route.

    \item \textbf{Pre-cutoff vs.\ post-cutoff contamination test per
    LLM baseline.}

    \item \textbf{Detailed baseline cards.} Prompts, retry policy,
    decoding (greedy / top-$p$ / temperature), tool list, max calls.

    \item \textbf{Pseudocode in appendix.} Projector, dynamic
    re-encoding, GRPO recipe, reward decomposition, tool-call protocol.

    \item \textbf{Decoding-setting rows on every baseline table.}

    \item \textbf{Single-repo release.} Code + checkpoints +
    LLAMIA-Bench harness + frozen eval script; anonymized URL for
    review.

    \item \textbf{Citations to add.}
    Coconut (Hao et al.), Maia-2,
    \cite{toolrl2025} (Re-Tool), \cite{latentmas2025},
    \cite{TNZse5q2Tr-creative-puzzles}, \cite{yRxX01oRIi-cot-hurts},
    \cite{Leela-lookahead-8zg9sO4ttV}, \cite{chessformer-2ltBRzEHyd},
    \cite{Allie-bc2H72hGxB}, \cite{DiffuSearch-A9y3LFX4ds},
    \cite{afaakBqkvb-obscuro}, \cite{CvjXlsBLCX-strength-est},
    \cite{kefUcn22mk-chessqa}, \cite{LLM-CHESS-65R1Dbfwzk},
    \cite{lnIlH0hfek-mom}, \cite{Maia4All-bwhLqFjsxd},
    \cite{cmN8Wbvanr-card-games}, \cite{BLOIB8CwBI-soundness},
    \cite{gBz9NMbvYS-maia2-sae}, \cite{Wxl0JMgDoU-maia2-sae},
    \cite{TIGQIem1na-comma}, Karvonen et al.\ 2024.

    \item \textbf{Re-render every table with $\leq 4$ decimal places
    and CIs on every numeric cell.}

    \item \textbf{Symbol-overload fixes.} Probe $e$ vs.\ Elo $e$;
    $\mathcal{B}$ vs.\ $B$; etc.

    \item \textbf{Backbone-refresh implementation.} Move from
    Vicuna-13B (2023-vintage) to Qwen3-13B-Instruct or Llama-4 8B/70B.
    \emph{Purpose:} reviewer-defensive on backbone freshness;
    Vicuna-13B becomes a single legacy ablation row.
\end{itemize}

\subsubsection*{P1}

\begin{itemize}[leftmargin=*]
    \item \textbf{Strict no-retry greedy gameplay protocol.}
    \cite{lnIlH0hfek-mom}-style.
    \emph{Purpose:} eliminates ``you got 7 retries per move.''

    \item \textbf{Full SAE feature catalog at anonymized URL.}

\end{itemize}

% =============================================================================
\subsection{Risks and contingencies}
\label{sec:todo_risks}
% =============================================================================

\begin{itemize}[leftmargin=*]
    \item \textbf{LLAMIA loses to Group~C on D1 / single-step tasks.}
    Likely. \emph{Plan:} lean on D2/D3 sustained-interaction metrics
    where re-encoding makes a structural difference; pre-write the
    ``parity at D1, win at D2/D3'' narrative.

    \item \textbf{LLAMIA loses to Group~A (frontier reasoning) at any
    spoke.} Possible (e.g.\ Claude Opus 4.7 on commentary).
    \emph{Plan:} report honestly; lean on the matched-recipe Group~F
    comparison for the structural claim.

    \item \textbf{The counter-intuitive prediction does not hold} --
    Group~B with engine access does outperform Group~A.
    \emph{Plan:} the verbalization bottleneck is weaker than thought;
    re-frame the contribution as ``latent internalization is more
    \emph{efficient} than tool-calling at matched performance''
    (Communication Efficiency / CE becomes the headline).

    \item \textbf{SAE features on the LLM-side projection are not
    interpretable.}
    \emph{Plan:} dual SAE (agent-side $\vh_s$ and LLM-side $\vz_s$);
    report alignment; lean on causal interventions which only need
    \emph{some} identifiable directions; frame as existence proof.

    \item \textbf{LLM-size $\times$ agent-strength scaling is flat in
    our range.}
    \emph{Plan:} show the regime where it breaks (very weak LLM or
    agent) and frame qualitatively as ``trends'' not ``laws.''

    \item \textbf{InvokeAgent impact is prompt-sensitive.}
    \emph{Plan:} three ICP prompt templates per theme; report variance.

    \item \textbf{GRPO entropy collapse.}
    \emph{Plan:} diversity filtering per
    \cite{TNZse5q2Tr-creative-puzzles}; cite Wang/Cui/Yu 2025 (DAPO).

    \item \textbf{Lc0 self-play games leak into LLAMIA-Bench.}
    \emph{Plan:} the data-leakage audit table catches this.

    \item \textbf{BC accuracy gains read cosmetic.}
    \emph{Plan:} every \% delta anchored to an Elo-equivalent via
    \cite{CvjXlsBLCX-strength-est}.
\end{itemize}

% =============================================================================
\subsection{Reference briefs}
\label{sec:todo_briefs}
% =============================================================================

Reference material for the writing tasks above. Each brief is a spec,
not a paper section; writing tasks point here.

\subsubsection*{Brief: LLAMIA-Bench positioning}
\label{brief:bench}

\paragraph{What LLAMIA-Bench is.}
A comprehensive suite of tasks for the structured evaluation of
\emph{LLM--specialist-agent collaboration}: the joint behaviour that
emerges when a generalist language model must perceive, calibrate,
explain, plan, imitate, steer, and remain robust under instruction
while paired with a domain specialist that thinks in a rich
continuous representation. Each task isolates one such collaborative
phenomenon and grades it against externally verifiable ground truth.
The benchmark's purpose is to map the surface of collaboration, not
to prosecute any single hypothesis about how the LM and the
specialist should be wired together.

\paragraph{One-line positioning.}
\emph{LLAMIA-Bench is the broadest ground-truth-anchored suite for
evaluating LLM--specialist-agent collaboration; chess is the only
present-day environment that supports its full breadth.}

\paragraph{Why chess (and only chess).}
The seven collaboration facets below all need a single environment to
be measured \emph{jointly} -- otherwise cross-facet comparisons
collapse into apples-to-oranges. Chess is the only domain that
satisfies all of the following at once:
\begin{itemize}[leftmargin=*]
    \item \textbf{Specialists with continuous, demonstrably superhuman
    representations.} Lc0/Stockfish produce policy and value tensors
    that an LM can be quantitatively shown to exploit; the
    representation is rich enough that ``reading the specialist'' is
    a non-trivial skill.
    \item \textbf{Stratified ground truth at every granularity.}
    Per-move (engine evaluation, near-oracle), per-turn (centipawn
    loss), per-game (W/D/L), per-player (Elo), per-cohort (skill
    bins), per-style (per-GM corpus), per-explanation (annotated game
    databases). Every collaboration facet has its own external
    rubric.
    \item \textbf{Calibrated, dense skill stratification.} Elo is
    standardised; cohorts down to ${\sim}100$-Elo bins are populated;
    behaviour at each bin is documented. Skill-conditioned imitation
    and persona-bending are first-class evaluable tasks
    \emph{only} here.
    \item \textbf{Decades of human--engine collaboration as a
    normative reference.} Centaur chess, opening preparation,
    engine-assisted commentary -- effective collaborative patterns
    are \emph{known}, so we can ask whether LM--agent systems
    recover them rather than inventing the rubric ourselves.
    \item \textbf{Fully observable but combinatorially deep.}
    Perception noise is removed as a confound; the specialist's
    advantage over the LM is purely strategic, isolating the
    collaboration interface as the variable of interest.
    \item \textbf{Multi-aspect single-state semantics.} One position
    simultaneously admits tactical, positional, strategic,
    opponent-modelling, and time-management readings -- multi-aspect
    explanation has nontrivial structure even on a single board.
    \item \textbf{Instruction-conditioned planning via Elo-swap.}
    A second player, configurable skill imbalance, and board-detectable
    strategic conditions combine into a clean instruction-conditioned
    planning testbed: win-rate is the direct metric, no auxiliary judge
    required.
    \item \textbf{Cheap, reproducible rollouts.} No real-world cost,
    sub-millisecond engine moves, round-robin tournaments at scale.
\end{itemize}
Adjacent domains cover only fragments. Medical imaging has ground
truth but no skill cohorts and no sustained adversarial play; coding
has no calibrated rating and only sparse explanation rubrics;
robotics introduces partial observability and perception noise as
confounds; multi-agent simulation lacks superhuman specialists. The
\emph{framework} -- evaluating collaboration along these facets --
is portable; the \emph{benchmark}, today, is chess by necessity.

\paragraph{The seven collaboration facets (organising axis).}
Every task in the suite operationalises exactly one facet:
\begin{itemize}[leftmargin=*]
    \item \textbf{Perception.} Does the LM see what the specialist
    sees? \textsc{Move-Class}, \textsc{Eval-1pos}, \textsc{Diff-1pos},
    \textsc{BC-1pos}.
    \item \textbf{Calibration.} Does the LM convey the specialist's
    confidence and strength faithfully? \textsc{Eval-Rat},
    \textsc{Diff-Rat}, \textsc{BC-Elo}.
    \item \textbf{Explanation.} Can the LM produce a multi-aspect
    rationale that is grounded in the specialist's state and useful
    to a human consumer? \textsc{Annot-Move}, \textsc{Int-Rat},
    \textsc{Comm-Move}.
    \item \textbf{Sustained planning.} Does the joint system stay
    coherent across an evolving sequence of states? \textsc{Comm-Game},
    \textsc{BC-Game}, \textsc{ICP-Win}.
    \item \textbf{Style and identity imitation.} Can the LM bend the
    specialist toward a target persona? \textsc{BC-Maia},
    \textsc{BC-GM}, \textsc{Stylo-ID}.
    \item \textbf{Instruction-conditioned adaptation.} Can a
    natural-language strategic instruction reshape the joint policy?
    \textsc{ICP-Win}, \textsc{ICP-Theme}.
\end{itemize}
Tasks carry a secondary tag for interaction depth
(\textsc{D1}/\textsc{D2}/\textsc{D3}) used by analyses that look
across facets; depth is a property of tasks, not the primary
organising axis of the benchmark.

\paragraph{Public benchmarks (separate, orthogonal).}
Maia-1900 BC, ChessBench Elo, Lichess solve rate, ChessQA, HPDv2 and
the cross-domain suites used in the persuasion / image-preference
section. They \emph{anchor} our systems on shared external scales
but they do not span the seven facets jointly; LLAMIA-Bench
complements them, it does not replace them.

\paragraph{Inclusion rule.}
A task enters LLAMIA-Bench iff (i) it operationalises exactly one of
the seven facets with externally verifiable ground truth, and (ii)
its inputs and outputs are identical across integration regimes (so
the same task can be re-run with text-only, verbalized-tool, or
latent-internalised access). Tasks failing (i) move to the public
anchor table; tasks failing (ii) are out of scope.

\paragraph{Integration interface as a corollary, not the design
goal.} The bench is built to evaluate collaboration; integration
regimes are evaluated \emph{through} it. Because every task has the
same I/O surface across regimes, the same suite can be re-run with
the integration mode swapped. The finding -- that interfaces degrade
differentially across the seven facets, with the verbalisation
penalty growing on sustained planning, instruction-conditioned
adaptation, and adversarial robustness -- emerges as a corollary of
running an existing comprehensive suite, not as the bench's purpose.
We accordingly report it as a finding (with the metrics in
\ref{brief:metrics} and the failure-mode taxonomy in
\ref{brief:failure}), not as the suite's reason for being.

\paragraph{Falsifiable, counter-intuitive prediction (a corollary).}
On the sustained-planning, instruction-conditioning, and adversarial
facets, frontier reasoning models (Claude Opus 4.7, GPT-5,
Gemini 2.5 Pro, DeepSeek-R1) augmented with the strongest available
chess engine through verbalized tools score no better than the same
models without engine access; latent internalisation at the same
scale wins the gap. If verbalisation is the bottleneck, a stronger
\emph{engine} cannot fix it -- only a different \emph{interface}
can. Falsified if Group~B saturates these facets; equally
informative either way.

\paragraph{Real-world relevance.}
The seven facets recur wherever a generalist LM collaborates with a
domain specialist (medical decision support, scientific simulation
steering, embodied control, security analysis, code review against
a static analyser). Today, chess is the only environment in which
all seven can be measured rigorously and jointly; the framework, not
the domain, ports.

\paragraph{What this benchmark is not.}
Not a hard-task showcase; not a low-data argument; not a chess
paper; not a competition; not a leaderboard; not an integration-mode
discriminator (that is a downstream use of the same data).

\paragraph{Acknowledged limitations.}
LM-judge dependence on explanation tasks (calibrated against a
human-rated subset; rule-based scoring used where feasible);
single-engine-class dependence (replicated across Lc0 nets and
Stockfish~17 in Group~G); chess-player human raters drawn from
community programmes rather than the authors; the ``why chess''
claim is sufficiency, not necessity -- the framework should port,
the benchmark itself does not.

\paragraph{Why this is novel.}
Existing chess benchmarks (ChessQA, ChessBench, LLM-CHESS) score
\emph{whether} an LLM solves a task; \cite{TIGQIem1na-comma} (COMMA)
treats inter-LM communication under information asymmetry as the
controlled variable in a multi-agent setting. LLAMIA-Bench is the
first to (i) span the seven LM--specialist collaboration facets
jointly under one environment, (ii) carry external ground truth at
every facet, and (iii) make integration-interface findings
recoverable as corollaries from a single comprehensive suite rather
than via purpose-built per-finding tasks.

\paragraph{Minimum-viable bench.}
One task per facet -- \textsc{Move-Class} (perception),
\textsc{Eval-Rat} (calibration), \textsc{Annot-Move} (explanation),
\textsc{Comm-Game} (sustained planning), \textsc{BC-Maia} (style and
identity), \textsc{ICP-Win} (instruction-conditioned adaptation). Six tasks, full
facet coverage, one radar.

\subsubsection*{Brief: thesis-aligned custom metrics}
\label{brief:metrics}

\begin{itemize}[leftmargin=*]
    \item \textbf{Verbalization Gap (VG).}
    $\mathrm{VG}(T) = s(\textsc{LLAMIA-full}, T) -
    s(\textsc{Tool-RL}, T)$. Per-task headline.
    \item \textbf{Depth-Compounding Slope (DCS).}
    Linear regression of $\mathrm{VG}(T)$ over depth bin
    $\{D1{=}1, D2{=}2, D3{=}3\}$ across all tasks. Bootstrap CIs over
    tasks. \emph{One scalar summarizes the bench.} Positive slope =
    thesis confirmed.
    \item \textbf{Re-encoding Yield (RY).}
    $\mathrm{RY}(T) = s(\textsc{LLAMIA-full}, T) -
    s(\textsc{LLAMIA-static}, T)$. Within-system contribution of
    dynamic re-encoding.
    \item \textbf{Communication Efficiency (CE).}
    Score / (LLM tokens + InvokeAgent calls + tool-call tokens),
    adapted from \cite{TIGQIem1na-comma}'s efficiency score.
\end{itemize}

\subsubsection*{Brief: failure-mode taxonomy}
\label{brief:failure}

\begin{itemize}[leftmargin=*]
    \item \textbf{F1 -- Verbalization clipping.} Agent's policy
    distribution rendered as ``top-$k$ with numbers,'' losing tail
    signal.
    \item \textbf{F2 -- Re-encoding lag.} LLM acts on a stale
    paraphrase after the state has evolved.
    \item \textbf{F3 -- Tool-call thrashing.} LLM repeatedly queries
    on near-identical positions.
    \item \textbf{F4 -- Plan abandonment.} LLM forgets a strategic
    instruction across a sequence of moves.
    \item \textbf{F5 -- Hallucinated state.} LLM references pieces,
    moves, or evaluations not present in the actual position.
    \item \textbf{F6 -- Honest blunder.} LLM and agent agreed on a bad
    plan -- the only legitimate failure mode.
\end{itemize}
Argument: \textbf{F1--F5 are exclusively verbalization failures;
LLAMIA structurally eliminates them; only F6 remains.}

\subsubsection*{Brief: methodology re-frame (SALT(SFT) + SALT(RL))}
\label{brief:methodology}

\textbf{Phase 1 -- SALT(SFT).} Stage 1 (projector alignment on
abundant synthetic data, LLM frozen) + Stage 2 (interleaved instruction
tuning). Warm-start; no longer the headline.

\textbf{Phase 2 -- SALT(RL).} GRPO (or DAPO) over rollouts that
include latent tokens and \texttt{<InvokeAgent>} calls. Reward
decomposed as: outcome reward (game result, BC accuracy, ICP win-rate,
calibrated judge-score on commentary), format reward (legal moves,
valid invocations), \emph{collaboration-shape reward} (encourages
invocation at decision-relevant moments). Diversity filtering against
entropy collapse.

\textbf{Backbone.} Qwen3-13B-Instruct or Llama-4 8B/70B (decided by
inference economics + InvokeAgent compatibility). Vicuna-13B retained
as a single legacy ablation row.

\subsubsection*{Brief: baseline grid}
\label{brief:baselines}

Every system carries an explicit access-mode tag and a recipe tag.

\textbf{Group A -- Frontier text-only.} Claude Opus 4.7, GPT-5/o4,
Gemini 2.5 Pro, DeepSeek-R1, Qwen3-235B-thinking. Access-mode
\texttt{text}; recipe \texttt{frontier-zero-shot}. Capability anchor.

\textbf{Group B -- Frontier with verbalized tools.} Same models with
ReAct-style access to Lc0-BT4 \texttt{/policy} and \texttt{/analyze},
Stockfish~17 \texttt{/analyze}. Access-mode \texttt{verbalized-tool};
recipe \texttt{frontier-react}. Standard tool-calling baseline.

\textbf{Group C -- RL-trained tool-using LLMs (the strawman).}
Re-Tool, ToolRL, Search-R1, AgentR1 applied to the chess tools, on
the same backbone as LLAMIA, matched compute. Access-mode
\texttt{verbalized-tool-rl}. The thesis-defining comparison.

\textbf{Group D -- Latent-CoT / continuous-reasoning.}
Coconut, latentmas, 2025--2026 implicit-reasoning baselines.

\textbf{Group E -- Domain specialists (per-task ceilings).}
\textsc{BC-Maia} = Maia-2; \textsc{BC-Time}/\textsc{BC-Elo} = Allie;
\textsc{BC-GM} = best-of-cohort per-GM model; gameplay = Lc0-BT4 /
Stockfish~17 / ChessFormer; \textsc{Diff-Rat}/\textsc{Int-Rat} =
generative-chess predictors (no rationale baseline -- the gap is the
contribution); \textsc{Comm-Move} = SCC; \textsc{Comm-Game} = none
(we introduce the task).

\textbf{[P1, appendix] Group F -- Matched-recipe LLAMIA ablations.}
\texttt{LLAMIA-text}, \texttt{LLAMIA-tool}, \texttt{LLAMIA-static},
\texttt{LLAMIA-full}. Same backbone, Stage-2 data, RL recipe; only
the projection differs. Appendix table; brief mention in main text only.

\textbf{Group G -- Lc0 architecture cross-check.}
LLAMIA with Lc0-BT4 / T60 / smaller Lc0 nets (varied depth/width).

\textbf{Decoding-protocol matrix.} Every baseline reported under
greedy / strict-no-retry and self-consistency (5 samples).

\textbf{Compute parity.} Every frontier baseline gets the same
per-task token budget as LLAMIA; LLAMIA's latent tokens count as
text-equivalent in the ``effective context'' column.

\subsubsection*{Brief: headline radar figure spec}
\label{brief:radar}

Single radar; spokes grouped around the circle in seven contiguous
arcs by collaboration facet (perception, calibration, explanation,
sustained planning, style/identity, instruction-conditioned
adaptation, adversarial robustness), in that order. Arc background
labels the facet; depth tag (D1/D2/D3) printed under each spoke
label as a secondary annotation. Three systems overlaid as connected
polygons: \texttt{LLAMIA-full}, the strongest Group~C
verbalized-tool-RL baseline, the strongest Group~A frontier reasoning
model. Per-spoke specialist ceilings drawn as separate dots, not a
polygon. Scores normalised per spoke to $[0, 1]$ via specialist
ceiling and text-only zero-shot floor; bounds reported alongside.
Dotted $\pm 95\%$ bootstrap CI bands. Caption: one-phrase definition
per spoke pointing to \texttt{appendix/dataset.tex}; DCS scalar (the
integration-interface corollary) with CI inset.

\paragraph{Status.}
This document is the active TODO. Items migrate to the relevant
appendix sections (\texttt{ablations.tex}, \texttt{study.tex},
\texttt{details.tex}, \texttt{dataset.tex}, \texttt{extension.tex})
as experiments land; pointers back here are removed once integrated.
