\section{Related Work}

\textbf{Text-mediated LLM-Agent Integration.}
The standard integration paradigm routes all communication between an LLM and a domain agent through language. ReAct~\cite{yao2022react} interleaves reasoning and action tokens within a single autoregressive sequence; Toolformer~\cite{schick2023toolformerlanguagemodelsteach} trains models to invoke external APIs and condition on their textual returns. Multi-agent pipelines such as HuggingGPT~\cite{shen2024hugginggpt} and AutoGen~\cite{wu2023autogen} extend this pattern: specialized agents deliver natural-language summaries to an LLM coordinator that reasons over them. Affordance-grounded variants (SayCan~\cite{ahn2022can}) and self-reflective loops (Reflexion~\cite{shinn2023reflexionlanguageagentsverbal}) refine coordination but preserve the same verbalization interface. Surveys of tool-augmented models~\cite{mialon2023augmented,ma2024sciagenttoolaugmentedlanguagemodels} confirm this is the dominant design across domains from generative simulation~\cite{sun2024factorsim} to autonomous driving~\cite{hu2024agentscodriver}. When both agents are language models, verbalization is lossless. When the domain agent is a pretrained neural network—a chess engine, a robotics policy, a game-playing agent—verbalization compresses its policy distribution and value landscape into a handful of tokens. Our work addresses precisely this information loss.

\textbf{Latent Reasoning in LLMs.}
A parallel line of work treats the continuous hidden states of a single LLM as a reasoning medium in their own right. CoCoNut~\cite{hao2024training} replaces token-level chain-of-thought with latent recurrent computation, demonstrating that hidden states encode structured semantic and procedural information that discrete token sequences cannot fully express~\cite{chen2025reasoning,zhang2025soft}. Representation engineering (RepE~\cite{zou2023representation}) and layer-of-thought (LoT~\cite{fungwacharakorn2024layer}) directly manipulate internal activations to steer downstream generation without explicit rationales. Recent work on latent collaboration~\cite{zheng2025thought,fu2025cache} passes KV-cache states between models to avoid serialization overhead. These methods confirm that representation-level communication can recover information lost at the token boundary. However, all existing approaches operate within a homogeneous population of language models; none addresses how to import representations from a \emph{heterogeneous}, independently pretrained non-language agent into an LLM's chain of thought.

\textbf{Chess Agents and Representation Analysis.}
Chess engines have evolved from handcrafted evaluation functions to neural architectures trained on billions of positions. Stockfish~\cite{maharaj2022chess} uses NNUE-style encoders~\cite{nasu2018efficiently}; AlphaZero~\cite{silver2017mastering} and Leela Chess Zero (LC0)~\cite{monroe2024mastering} encode board understanding in deep residual networks. Interpretability studies show that the encoder backbone \(G_\psi: S \to \mathbb{R}^d\) captures long-range planning signals, piece-value geometry, and strategic features that are not recoverable from a single best-move prediction~\cite{jenner2024evidence}. Transformer-based agents~\cite{ruoss2024grandmaster,monroe2024mastering} achieve grandmaster-level play without search, suggesting these representations encode information sufficient for multi-step lookahead. Prior LLM-chess work either trains dedicated models per task~\cite{jhamtani-etal-2018-learning,mcilroy2020aligning} or grounds an LLM via verbalized engine outputs~\cite{feng2024chessgpt}; neither exposes the engine's internal state to the LLM's reasoning process, which is the gap LLAMIA fills.
